---
title: "Score vs Wald test: multi-scenario comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Score vs Wald test: multi-scenario comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message=FALSE, warning=FALSE}
library(gsDesignNB)
library(data.table)
library(ggplot2)
library(gt)
```

## Introduction

This vignette compares the Wald and score test implementations in `mutze_test()` across a
factorial grid of negative binomial trial scenarios. We vary the control event rate
($\lambda_1$), overdispersion ($k$), and minimum inter-event gap to assess:

- **Type I error control** under $H_0: RR = 1$
- **Power** under $H_1: RR = 0.70$
- **Z-score distributions** to check asymptotic normality

Results are pre-computed by `data-raw/generate_score_sweep.R` and loaded here.

## Load pre-computed results

```{r}
results_file <- system.file("extdata", "score_sweep_results.rds",
                            package = "gsDesignNB")
if (results_file == "" && file.exists("../inst/extdata/score_sweep_results.rds")) {
  results_file <- "../inst/extdata/score_sweep_results.rds"
}
res <- readRDS(results_file)
config <- res$config
scenarios <- res$scenarios
```

```{r}
cat(sprintf(
  "Scenarios: %d | Power sims: %s | Null sims: %s | RR: %.2f | alpha: %.3f\n",
  nrow(scenarios),
  format(config$n_sims_power, big.mark = ","),
  format(config$n_sims_null, big.mark = ","),
  config$rr_power,
  config$alpha
))
```

### Scenario grid

```{r}
scenarios |>
  as.data.table() |>
  _[, .(lambda1, k, gap_days, n_total)] |>
  gt() |>
  tab_header(title = "Scenario grid with computed sample sizes") |>
  cols_label(
    lambda1 = "Control rate",
    k = "Dispersion (k)",
    gap_days = "Event gap (days)",
    n_total = "Total N"
  ) |>
  fmt_number(columns = lambda1, decimals = 2) |>
  fmt_number(columns = n_total, decimals = 0)
```

## Type I error comparison

```{r}
null_dt <- as.data.table(res$null_summary)
null_long <- melt(
  null_dt,
  id.vars = c("lambda1", "k", "gap_days", "n_total"),
  measure.vars = c("rate_wald", "rate_score"),
  variable.name = "test",
  value.name = "rejection_rate"
)
null_long[, test := fifelse(test == "rate_wald", "Wald", "Score")]

# Add SE
se_long <- melt(
  null_dt,
  id.vars = c("lambda1", "k", "gap_days"),
  measure.vars = c("se_wald", "se_score"),
  variable.name = "test",
  value.name = "se"
)
se_long[, test := fifelse(test == "se_wald", "Wald", "Score")]
null_long <- merge(null_long, se_long,
                   by = c("lambda1", "k", "gap_days", "test"))
```

```{r}
null_long |>
  _[, .(lambda1, k, gap_days, n_total, test, rejection_rate, se)] |>
  gt() |>
  tab_header(
    title = "Type I error rate by scenario",
    subtitle = sprintf("Nominal alpha = %.3f, %s null sims per scenario",
                        config$alpha, format(config$n_sims_null, big.mark = ","))
  ) |>
  cols_label(
    lambda1 = "Control rate",
    k = "Dispersion",
    gap_days = "Gap (days)",
    n_total = "N",
    test = "Test",
    rejection_rate = "Type I error",
    se = "SE"
  ) |>
  fmt_number(columns = c(rejection_rate, se), decimals = 4) |>
  fmt_number(columns = lambda1, decimals = 2)
```

```{r, fig.width=7, fig.height=5}
ggplot(null_long, aes(x = interaction(lambda1, k, sep = " / "),
                      y = rejection_rate, color = test, shape = test)) +

  geom_point(size = 3, position = position_dodge(width = 0.3)) +
  geom_errorbar(aes(ymin = rejection_rate - 1.96 * se,
                     ymax = rejection_rate + 1.96 * se),
                width = 0.15, position = position_dodge(width = 0.3)) +
  geom_hline(yintercept = config$alpha, linetype = "dashed", color = "grey40") +
  facet_wrap(~ paste0("Gap = ", gap_days, "d")) +
  labs(
    title = "Type I error: Wald vs Score test",
    x = expression(lambda[1] ~ "/" ~ k),
    y = "Rejection rate",
    color = "Test", shape = "Test"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Power comparison

```{r}
power_dt <- as.data.table(res$power_summary)
power_long <- melt(
  power_dt,
  id.vars = c("lambda1", "k", "gap_days", "n_total"),
  measure.vars = c("rate_wald", "rate_score"),
  variable.name = "test",
  value.name = "power"
)
power_long[, test := fifelse(test == "rate_wald", "Wald", "Score")]

se_power <- melt(
  power_dt,
  id.vars = c("lambda1", "k", "gap_days"),
  measure.vars = c("se_wald", "se_score"),
  variable.name = "test",
  value.name = "se"
)
se_power[, test := fifelse(test == "se_wald", "Wald", "Score")]
power_long <- merge(power_long, se_power,
                    by = c("lambda1", "k", "gap_days", "test"))
```

```{r}
power_long |>
  _[, .(lambda1, k, gap_days, n_total, test, power, se)] |>
  gt() |>
  tab_header(
    title = "Power by scenario",
    subtitle = sprintf("RR = %.2f, %s power sims per scenario",
                        config$rr_power, format(config$n_sims_power, big.mark = ","))
  ) |>
  cols_label(
    lambda1 = "Control rate",
    k = "Dispersion",
    gap_days = "Gap (days)",
    n_total = "N",
    test = "Test",
    power = "Power",
    se = "SE"
  ) |>
  fmt_number(columns = c(power, se), decimals = 4) |>
  fmt_number(columns = lambda1, decimals = 2)
```

```{r, fig.width=7, fig.height=5}
ggplot(power_long, aes(x = interaction(lambda1, k, sep = " / "),
                       y = power, color = test, shape = test)) +
  geom_point(size = 3, position = position_dodge(width = 0.3)) +
  geom_errorbar(aes(ymin = power - 1.96 * se,
                     ymax = power + 1.96 * se),
                width = 0.15, position = position_dodge(width = 0.3)) +
  geom_hline(yintercept = config$power_target, linetype = "dashed", color = "grey40") +
  facet_wrap(~ paste0("Gap = ", gap_days, "d")) +
  labs(
    title = "Power: Wald vs Score test",
    subtitle = sprintf("Target = %.0f%%, RR = %.2f",
                        100 * config$power_target, config$rr_power),
    x = expression(lambda[1] ~ "/" ~ k),
    y = "Power",
    color = "Test", shape = "Test"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Z-score density comparison (null simulations)

Under $H_0$, the Z-statistics should follow $N(0, 1)$ if the asymptotic approximation holds.

```{r, fig.width=8, fig.height=6}
z_null <- as.data.table(res$z_sample_null)
sc_info <- data.table(
  scenario_id = seq_len(nrow(scenarios)),
  scenarios[, c("lambda1", "k", "gap_days")]
)
z_null <- merge(z_null, sc_info, by = "scenario_id")
z_null[, label := sprintf("l1=%.2f k=%.1f gap=%dd", lambda1, k, gap_days)]

z_null_long <- melt(
  z_null,
  id.vars = c("scenario_id", "label"),
  measure.vars = c("z_wald", "z_score"),
  variable.name = "test",
  value.name = "z"
)
z_null_long[, test := fifelse(test == "z_wald", "Wald", "Score")]
z_null_long <- z_null_long[is.finite(z)]

ggplot(z_null_long, aes(x = z, color = test)) +
  geom_density(linewidth = 0.6) +
  stat_function(fun = dnorm, color = "black", linetype = "dashed", linewidth = 0.4) +
  facet_wrap(~ label, scales = "free_y") +
  labs(
    title = "Null Z-score densities: Wald vs Score",
    subtitle = "Dashed line = N(0,1) reference",
    x = "Z-statistic", y = "Density",
    color = "Test"
  ) +
  theme_minimal() +
  coord_cartesian(xlim = c(-4, 4))
```

## Fallback method frequency

When the negative binomial MLE fails to converge or yields non-overdispersed estimates,
`mutze_test()` falls back to Poisson or method-of-moments estimation.

```{r}
null_fb <- as.data.table(res$null_summary)
fb_cols <- c("lambda1", "k", "gap_days",
             "pct_fallback_poisson_wald", "pct_fallback_mom_wald",
             "pct_fallback_poisson_score", "pct_fallback_mom_score")
null_fb[, ..fb_cols] |>
  gt() |>
  tab_header(title = "Fallback method frequency (%, null sims)") |>
  cols_label(
    lambda1 = "Control rate",
    k = "Dispersion",
    gap_days = "Gap (days)",
    pct_fallback_poisson_wald = "Poisson (Wald)",
    pct_fallback_mom_wald = "MoM (Wald)",
    pct_fallback_poisson_score = "Poisson (Score)",
    pct_fallback_mom_score = "MoM (Score)"
  ) |>
  fmt_number(columns = starts_with("pct_"), decimals = 1) |>
  fmt_number(columns = lambda1, decimals = 2)
```

## Summary

- Both Wald and score tests control Type I error near the nominal $\alpha = `r config$alpha`$
  across all scenarios examined.
- Power is similar between the two tests, as expected for well-powered designs.
- The score test tends to be slightly more conservative (lower Type I error) in
  scenarios with smaller sample sizes or lower event rates.
- Z-score distributions under the null closely follow $N(0,1)$, confirming the
  validity of the asymptotic approximation for the sample sizes used.
