---
title: "Score vs Wald test: multi-scenario comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Score vs Wald test: multi-scenario comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message=FALSE, warning=FALSE}
library(gsDesignNB)
library(data.table)
library(ggplot2)
library(plotly)
library(DT)
library(crosstalk)
```

## Introduction

This vignette compares the Wald and score test implementations in `mutze_test()` across a
factorial grid of negative binomial trial scenarios. We also compare sample sizes computed
using the **Wald variance formula** ($V_1$ only) and the **score variance formula**
($\sqrt{V_0}$ and $\sqrt{V_1}$) in `sample_size_nbinom()`.

The full $2 \times 2$ factorial comparison is:

| | Wald-sized trial | Score-sized trial |
|---|---|---|
| **Wald test** | Wald / Wald | Score / Wald |
| **Score test** | Wald / Score | Score / Score |

We assess:

- **Type I error control** under $H_0: RR = 1$
- **Power** under $H_1: RR = 0.70$
- **Z-score distributions** to check asymptotic normality

All tables are interactive: use the column filters to narrow results.
All figures are interactive: hover for values, click legend entries to toggle, and
use the toolbar to zoom/pan.

Results are pre-computed by `data-raw/generate_score_sweep.R` and loaded here.

## Load pre-computed results

```{r}
results_file <- system.file("extdata", "score_sweep_results.rds",
                            package = "gsDesignNB")
if (results_file == "" && file.exists("../inst/extdata/score_sweep_results.rds")) {
  results_file <- "../inst/extdata/score_sweep_results.rds"
}
res <- readRDS(results_file)
config <- res$config
scenarios <- as.data.table(res$scenarios)
base_grid <- as.data.table(res$base_grid)
```

```{r}
cat(sprintf(
  "Expanded scenarios: %d | Power sims: %s | Null sims: %s | RR: %.2f | alpha: %.3f\n",
  nrow(scenarios),
  format(config$n_sims_power, big.mark = ","),
  format(config$n_sims_null, big.mark = ","),
  config$rr_power,
  config$alpha
))
```

### Scenario grid

The base scenario grid varies control event rate ($\lambda_1$), overdispersion ($k$),
and minimum inter-event gap. For each base scenario, sample sizes are computed using
both the Wald and score variance formulas.

```{r}
base_display <- base_grid[, .(
  `Control rate` = lambda1,
  `Dispersion (k)` = k,
  `Event gap (days)` = gap_days,
  `N (Wald sizing)` = n_wald,
  `N (Score sizing)` = n_score,
  `Difference` = n_wald - n_score
)]

datatable(
  base_display,
  caption = "Base scenario grid with sample sizes by method",
  filter = "top",
  options = list(pageLength = 15, dom = "ftip"),
  rownames = FALSE
) |> formatRound(columns = "Control rate", digits = 2)
```

## Type I error comparison

```{r}
null_dt <- as.data.table(res$null_summary)
null_long <- melt(
  null_dt,
  id.vars = c("lambda1", "k", "gap_days", "n_total", "sizing"),
  measure.vars = c("rate_wald", "rate_score"),
  variable.name = "test",
  value.name = "rejection_rate"
)
null_long[, test := fifelse(test == "rate_wald", "Wald", "Score")]

se_long <- melt(
  null_dt,
  id.vars = c("lambda1", "k", "gap_days", "sizing"),
  measure.vars = c("se_wald", "se_score"),
  variable.name = "test",
  value.name = "se"
)
se_long[, test := fifelse(test == "se_wald", "Wald", "Score")]
null_long <- merge(null_long, se_long,
                   by = c("lambda1", "k", "gap_days", "sizing", "test"))
null_long[, combo := paste0(sizing, "-sized / ", test, " test")]
```

```{r}
null_display <- null_long[order(lambda1, k, gap_days, sizing, test),
  .(
    `Control rate` = lambda1,
    Dispersion = k,
    `Gap (days)` = gap_days,
    Sizing = sizing,
    N = n_total,
    Test = test,
    `Type I error` = round(rejection_rate, 4),
    SE = round(se, 4)
  )
]

datatable(
  null_display,
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Type I error rate — nominal alpha = %.3f, %s null sims/scenario",
            config$alpha, format(config$n_sims_null, big.mark = ","))
  ),
  filter = "top",
  options = list(pageLength = 20, dom = "ftip"),
  rownames = FALSE
)
```

```{r, fig.width=8, fig.height=6}
null_long[, scenario := paste0("λ₁=", lambda1, " k=", k)]

p_null <- ggplot(null_long,
  aes(x = scenario, y = rejection_rate,
      color = combo, shape = test,
      text = paste0("Sizing: ", sizing, "\nTest: ", test,
                    "\nRate: ", round(rejection_rate, 4),
                    "\nN: ", n_total))) +
  geom_point(size = 2.5, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = rejection_rate - 1.96 * se,
                     ymax = rejection_rate + 1.96 * se),
                width = 0.2, position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = config$alpha, linetype = "dashed", color = "grey40") +
  facet_wrap(~ paste0("Gap = ", gap_days, "d")) +
  labs(
    title = "Type I error: sizing method × test type",
    x = NULL, y = "Rejection rate",
    color = "Sizing / Test", shape = "Test"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(p_null, tooltip = "text") |>
  layout(legend = list(orientation = "v", x = 1.02, y = 0.5))
```

## Power comparison

```{r}
power_dt <- as.data.table(res$power_summary)
power_long <- melt(
  power_dt,
  id.vars = c("lambda1", "k", "gap_days", "n_total", "sizing"),
  measure.vars = c("rate_wald", "rate_score"),
  variable.name = "test",
  value.name = "power"
)
power_long[, test := fifelse(test == "rate_wald", "Wald", "Score")]

se_power <- melt(
  power_dt,
  id.vars = c("lambda1", "k", "gap_days", "sizing"),
  measure.vars = c("se_wald", "se_score"),
  variable.name = "test",
  value.name = "se"
)
se_power[, test := fifelse(test == "se_wald", "Wald", "Score")]
power_long <- merge(power_long, se_power,
                    by = c("lambda1", "k", "gap_days", "sizing", "test"))
power_long[, combo := paste0(sizing, "-sized / ", test, " test")]
```

```{r}
power_display <- power_long[order(lambda1, k, gap_days, sizing, test),
  .(
    `Control rate` = lambda1,
    Dispersion = k,
    `Gap (days)` = gap_days,
    Sizing = sizing,
    N = n_total,
    Test = test,
    Power = round(power, 4),
    SE = round(se, 4)
  )
]

datatable(
  power_display,
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    sprintf("Power — RR = %.2f, %s power sims/scenario",
            config$rr_power, format(config$n_sims_power, big.mark = ","))
  ),
  filter = "top",
  options = list(pageLength = 20, dom = "ftip"),
  rownames = FALSE
)
```

```{r, fig.width=8, fig.height=6}
power_long[, scenario := paste0("λ₁=", lambda1, " k=", k)]

p_power <- ggplot(power_long,
  aes(x = scenario, y = power,
      color = combo, shape = test,
      text = paste0("Sizing: ", sizing, "\nTest: ", test,
                    "\nPower: ", round(power, 4),
                    "\nN: ", n_total))) +
  geom_point(size = 2.5, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = power - 1.96 * se,
                     ymax = power + 1.96 * se),
                width = 0.2, position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = config$power_target, linetype = "dashed", color = "grey40") +
  facet_wrap(~ paste0("Gap = ", gap_days, "d")) +
  labs(
    title = "Power: sizing method × test type",
    subtitle = sprintf("Target = %.0f%%, RR = %.2f",
                        100 * config$power_target, config$rr_power),
    x = NULL, y = "Power",
    color = "Sizing / Test", shape = "Test"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(p_power, tooltip = "text") |>
  layout(legend = list(orientation = "v", x = 1.02, y = 0.5))
```

## Z-score density comparison (null simulations)

Under $H_0$, the Z-statistics should follow $N(0, 1)$ if the asymptotic approximation holds.

```{r, fig.width=9, fig.height=7}
z_null <- as.data.table(res$z_sample_null)
sc_info <- data.table(
  scenario_id = seq_len(nrow(scenarios)),
  scenarios[, .(lambda1, k, gap_days, sizing)]
)
z_null <- merge(z_null, sc_info, by = "scenario_id")
z_null[, label := sprintf("l1=%.2f k=%.1f gap=%dd (%s)",
                           lambda1, k, gap_days, sizing)]

z_null_long <- melt(
  z_null,
  id.vars = c("scenario_id", "label", "sizing"),
  measure.vars = c("z_wald", "z_score"),
  variable.name = "test",
  value.name = "z"
)
z_null_long[, test := fifelse(test == "z_wald", "Wald", "Score")]
z_null_long <- z_null_long[is.finite(z)]

p_z <- ggplot(z_null_long, aes(x = z, color = test)) +
  geom_density(linewidth = 0.6) +
  stat_function(fun = dnorm, color = "black", linetype = "dashed", linewidth = 0.4) +
  facet_wrap(~ label, scales = "free_y") +
  labs(
    title = "Null Z-score densities: Wald vs Score",
    subtitle = "Dashed line = N(0,1) reference",
    x = "Z-statistic", y = "Density",
    color = "Test"
  ) +
  theme_minimal() +
  coord_cartesian(xlim = c(-4, 4))

ggplotly(p_z) |>
  layout(legend = list(orientation = "h", x = 0.3, y = -0.05))
```

## Fallback method frequency

When the negative binomial MLE fails to converge or yields non-overdispersed estimates,
`mutze_test()` falls back to Poisson or method-of-moments estimation.

```{r}
null_fb <- as.data.table(res$null_summary)
fb_display <- null_fb[, .(
  `Control rate` = lambda1,
  Dispersion = k,
  `Gap (days)` = gap_days,
  Sizing = sizing,
  `Poisson (Wald)` = round(pct_fallback_poisson_wald, 1),
  `MoM (Wald)` = round(pct_fallback_mom_wald, 1),
  `Poisson (Score)` = round(pct_fallback_poisson_score, 1),
  `MoM (Score)` = round(pct_fallback_mom_score, 1)
)]

datatable(
  fb_display,
  caption = "Fallback method frequency (%, null sims)",
  filter = "top",
  options = list(pageLength = 20, dom = "ftip"),
  rownames = FALSE
)
```

## Summary

- Both Wald and score tests control Type I error near the nominal $\alpha = `r config$alpha`$
  across all scenarios and sizing methods.
- The **score-sized** trials use the two-variance formula
  $(z_\alpha \sqrt{V_0} + z_\beta \sqrt{V_1})^2 / \theta^2$,
  which typically yields slightly larger sample sizes than the Wald formula.
- Power is similar between test types for a given sizing; score-sized trials may show
  marginally higher power due to the larger sample size.
- The score test tends to be slightly more conservative (lower Type I error),
  especially in scenarios with smaller sample sizes or lower event rates.
- Z-score distributions under the null closely follow $N(0,1)$, confirming the
  validity of the asymptotic approximation for the sample sizes used.
