---
title: "Simulating recurrent events"
output: rmarkdown::html_vignette
bibliography: gsDesignNB.bib
vignette: >
  %\VignetteIndexEntry{Simulation of recurrent events}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message=FALSE, warning=FALSE}
library(gsDesignNB)
library(data.table)
library(ggplot2)
library(gt)
library(scales)
```

This vignette demonstrates how to use the `nb_sim()` function to simulate a 2-arm clinical trial with recurrent event outcomes.

## Simulation setup

We will simulate a small trial with the following parameters:

*   **Enrollment:** 20 patients total, recruited over 5 months with a constant rate.
*   **Treatments:** Two groups (Control vs. Experimental) with a 1:1 allocation ratio.
*   **Event Rates:**
    *   Control: 0.5 events per year.
    *   Experimental: 0.3 events per year.
*   **Dropout:**
    *   Control: 10% annual dropout rate.
    *   Experimental: 5% annual dropout rate.
*   **Maximum Follow-up:** Each patient is followed for a maximum of 2 years from randomization.

### Defining input parameters

```{r}
# Enrollment: rate of 4 patients/month for 5 months -> ~20 patients
enroll_rate <- data.frame(
  rate = 4,
  duration = 5
)

# Failure rates (events per unit time)
fail_rate <- data.frame(
  treatment = c("Control", "Experimental"),
  rate = c(0.5, 0.3) # events per year (assuming time unit is year, adjust enroll duration if needed)
)

# Let's ensure time units are consistent.
# If fail_rate is per year, then durations should be in years.
# 5 months = 5/12 years.
enroll_rate <- data.frame(
  rate = 20 / (5 / 12), # 20 patients over 5/12 years
  duration = 5 / 12
)

# Dropout rates (per year)
dropout_rate <- data.frame(
  treatment = c("Control", "Experimental"),
  rate = c(0.1, 0.05),
  duration = c(100, 100) # constant rate for long duration
)

# Maximum follow-up per patient (years)
max_followup <- 2
```

## Running the simulation

```{r}
set.seed(123)

sim_data <- nb_sim(
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  dropout_rate = dropout_rate,
  max_followup = max_followup,
  n = 20
)

head(sim_data)
```

The output contains multiple rows per subject:

*   `event = 1`: An actual recurrent event.
*   `event = 0`: The censoring time (either due to dropout or reaching `max_followup`).

## Analyzing the data

We can aggregate the data to calculate the observed event rates and total follow-up time for each group.

```{r}
sim_dt <- as.data.table(sim_data)
sim_dt[, censor_followup := ifelse(event == 0, tte, 0)]
summary_stats <- sim_dt[
  ,
  .(
    n_subjects = uniqueN(id),
    total_events = sum(event == 1),
    total_followup = sum(censor_followup),
    observed_rate = sum(event == 1) / sum(censor_followup)
  ),
  by = treatment
]
summary_stats |>
  gt() |>
  tab_header(title = "Summary Statistics by Treatment") |>
  cols_label(
    treatment = "Treatment",
    n_subjects = "N",
    total_events = "Events",
    total_followup = "Follow-up",
    observed_rate = "Rate"
  ) |>
  fmt_number(columns = total_followup, decimals = 2) |>
  fmt_number(columns = observed_rate, decimals = 3)
```

### Inspect first ten records

Before plotting, we can look at the first ten records from the simulated dataset.

```{r}
head(sim_data, 10)
```

### Plotting events

We can visualize the events and censoring times for each subject. To avoid any
side-effects from `data.table`, we convert the dataset back to a plain data frame.
We also display an event gap after each event where no new events can be recorded.
For illustration purposes in this plot, we show a **20-day gap** so it is visible on the timeline.

```{r, fig.width=7, fig.height=5, fig.alt="Patient timelines with events (dots), 20-day gaps (gray segments), and censoring (X)"}
sim_plot <- as.data.frame(sim_data)
names(sim_plot) <- make.names(names(sim_plot), unique = TRUE)
events_df <- sim_plot[sim_plot$event == 1, ]
censor_df <- sim_plot[sim_plot$event == 0, ]

# Define a 20-day gap for visualization
gap_duration <- 20 / 365.25

# Create segments for the gap after each event
events_df$gap_start <- events_df$tte
events_df$gap_end <- events_df$tte + gap_duration

ggplot(sim_plot, aes(x = tte, y = factor(id), color = treatment)) +
  geom_line(aes(group = id), color = "gray80") +
  # Add gap segments
  geom_segment(
    data = events_df,
    aes(x = gap_start, xend = gap_end, y = factor(id), yend = factor(id)),
    color = "gray50", linewidth = 2, alpha = 0.7
  ) +
  geom_point(data = events_df, shape = 19, size = 2) +
  geom_point(data = censor_df, shape = 4, size = 3) +
  labs(
    title = "Patient Timelines",
    x = "Time from Randomization (Years)",
    y = "Patient ID",
    caption = "Dots = Events, Gray Bars = 20-day Gap, X = Censoring/Dropout"
  ) +
  theme_minimal()
```

## Cutting data by analysis date

We can truncate the simulated data at an interim analysis date using
`cut_data_by_date()`. The function returns a single record per participant with
the truncated follow-up time (`tte`) and number of observed events.
By default, no gap (`event_gap = 0`) is applied after each event.
If a positive `event_gap` is used, no new events are counted during that gap,
and this time is excluded from time at risk.

```{r}
cut_summary <- cut_data_by_date(sim_data, cut_date = 1.5)
head(cut_summary)
```

## Missing data assumptions and imputation

The primary analysis in `gsDesignNB` is an observed-data rate analysis: each
participant contributes the recurrent events observed before the analysis cut and
the corresponding exposure time. In `mutze_test()`, this appears as a negative
binomial or Poisson log-rate model with `offset(log(tte))`; unobserved future
events after censoring are not filled in.

Under the current estimand and model, missing at random (MAR) censoring does not
by itself require multiple imputation (MI). Rubin's ignorability result states
that, for direct likelihood or Bayesian inference, the missing-data process can
be ignored when the data are MAR and the parameters governing missingness are
distinct from the outcome-model parameters [@rubin1976inference]. For this
package, the practical reading is:

*   if dropout or administrative censoring is independent of future unobserved
    event burden after conditioning on the variables and histories represented in
    the analysis model, use the observed events and observed exposure;
*   do not discard partially followed participants, since their observed events
    and exposure remain informative;
*   MI under MAR is optional as a modeling or reporting strategy, not a
    requirement for valid point estimates and standard errors from a correctly
    specified observed-data likelihood.

This statement is narrower than saying that any available-case analysis is valid
under MAR. If MAR depends on auxiliary covariates, prior event history, center,
season, adherence, or other observed predictors that are not represented in the
analysis, those variables should be added to the analysis, used in an inverse
probability weighting model, or used in a compatible imputation model. A
complete-case analysis of only subjects with full follow-up is generally a
different and less efficient analysis, and may be biased when incomplete
participants differ from completers.

For missing not at random (MNAR), the missing future event process is not
identified from observed data alone. Imputation can be useful, but only as part
of an explicit sensitivity analysis [@nrc2010missing]. A future MNAR MI routine
for this package should first preserve enough information to define what is
missing:

*   record the censoring reason, such as dropout, administrative cut, maximum
    follow-up, or analysis cut;
*   keep the planned follow-up horizon so the unobserved exposure window after
    dropout can be calculated;
*   include observed predictors of both dropout and event burden, such as
    treatment, observed events, observed exposure, season, enrollment time, and
    dropout reason.

An MNAR imputation model should then impute post-dropout recurrent-event counts
or event times from a negative binomial recurrent-event model with an exposure
offset. Sensitivity parameters can shift the post-dropout event rate, for
example by using treatment-specific multipliers
`\lambda_post = \lambda_model * exp(delta_g)`. Running a grid of `delta_g`
values gives a tipping-point or pattern-mixture sensitivity analysis. Each
imputed dataset should be analyzed with the same `mutze_test()` estimand and the
log-rate-ratio estimates and variances combined across imputations using Rubin
rules. If `event_gap` is important, imputing event times rather than only total
counts is preferable so at-risk time after events is handled consistently.

## Wald test [@mutze2019group]

Using the truncated data we can run a negative binomial Wald test as described by @mutze2019group.

```{r}
mutze_res <- mutze_test(cut_summary)
mutze_res
```

## Finding analysis date for target events

Often we want to perform an interim analysis not at a fixed calendar date, but
when a specific number of events have accumulated. We can find this date using
`get_analysis_date()` and then cut the data accordingly. This function also
respects the `event_gap` (defaulting to 5 days).

```{r}
# Target 15 total events
target_events <- 15
analysis_date <- get_analysis_date(sim_data, planned_events = target_events)

print(paste("Calendar date for", target_events, "events:", round(analysis_date, 3)))

# Cut data at this date
cut_events <- cut_data_by_date(sim_data, cut_date = analysis_date)

# Verify event count
sum(cut_events$events)
```

## Generating and verifying negative binomial data

The `nb_sim()` function can also generate data where the counts follow a negative binomial distribution. This is achieved by providing a `dispersion` parameter in the `fail_rate` data frame. The dispersion parameter $k$ relates the variance to the mean as $Var(Y) = \mu + k\mu^2$.

### Simulation with dispersion

We define a scenario with a known dispersion of 0.5.

```{r}
# Define failure rates with dispersion
fail_rate_nb <- data.frame(
  treatment = "Control",
  rate = 10, # Mean event rate
  dispersion = 2 # Variance = mean + 2 * mean^2
)

enroll_rate_nb <- data.frame(
  rate = 100,
  duration = 1
)

set.seed(1)
# Simulate 10000 subjects for a quick local review build
sim_nb <- nb_sim(
  enroll_rate = enroll_rate_nb,
  fail_rate = fail_rate_nb,
  max_followup = 1,
  n = 10000,
  block = "Control" # Assign all to Control for simplicity
)
```

### Verifying the dispersion parameter

We can verify that the simulated data reflects the input dispersion parameter by estimating it back from the data. 
We use the Method of Moments (MoM) estimator:

$$ \hat{k} = \frac{Var(Y) - \bar{Y}}{\bar{Y}^2} $$

```{r}
# Count events per subject
counts_nb <- as.data.table(sim_nb)[, .(events = sum(event)), by = id]

m <- mean(counts_nb$events)
v <- var(counts_nb$events)
k_mom <- (v - m) / (m^2)

# Theoretical values
mu_true <- 10
k_true <- 2
v_true <- mu_true + k_true * mu_true^2

# Also estimate using GLM
# We use MASS::glm.nb to fit the negative binomial model
# We suppress warnings because fitting intercept-only models on simulated data
# can occasionally produce convergence warnings despite valid estimates.
k_glm <- tryCatch(
  {
    fit <- suppressWarnings(MASS::glm.nb(events ~ 1, data = counts_nb))
    1 / fit$theta
  },
  error = function(e) NA
)

print(paste("True Mean:", mu_true, "| Observed Mean:", signif(m, 4)))
print(paste("True Variance:", v_true, "| Observed Variance:", signif(v, 4)))
print(paste("True Dispersion:", k_true))
print(paste("Estimated Dispersion (MoM):", signif(k_mom, 4)))
print(paste("Estimated Dispersion (GLM):", signif(k_glm, 4)))
```

### Visualizing the distribution

We can compare the observed distribution of event counts to the theoretical negative binomial distribution.

```{r}
# Calculate observed proportions
obs_dist <- counts_nb[, .N, by = events]
obs_dist[, prop := N / sum(N)]
obs_dist[, type := "Observed"]

# Calculate theoretical probabilities
# Parameters: mu = 10, size = 1/k = 1/2 = 0.5
mu <- 10
k <- 2
size <- 1/k
max_events <- max(obs_dist$events)
theo_probs <- dnbinom(0:max_events, size = size, mu = mu)
theo_dist <- data.table(
  events = 0:max_events,
  N = NA,
  prop = theo_probs,
  type = "Theoretical"
)

# Combine for plotting
plot_data <- rbind(obs_dist, theo_dist)

# Filter for visualization (limit x-axis to 50)
plot_data <- plot_data[events <= 50]

# Plot
ggplot(plot_data, aes(x = events, y = prop, fill = type)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
  scale_y_log10(labels = scales::label_number()) +
  labs(
    title = "Observed vs. Theoretical Negative Binomial Distribution",
    subtitle = paste("Mean =", mu, ", Dispersion =", k, "(Log Scale)"),
    x = "Number of Events",
    y = "Proportion",
    fill = "Distribution"
  ) +
  theme_minimal()
```

## References
