---
title: "Simulation of Recurrent Events"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation of Recurrent Events}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(nbDesign)
library(dplyr)
```

This vignette demonstrates how to use the `nb_sim()` function to simulate a 2-arm clinical trial with recurrent event outcomes.

## Simulation Setup

We will simulate a small trial with the following parameters:

*   **Enrollment:** 20 patients total, recruited over 5 months with a constant rate.
*   **Treatments:** Two groups (Control vs. Experimental) with a 1:1 allocation ratio.
*   **Event Rates:**
    *   Control: 0.5 events per year.
    *   Experimental: 0.3 events per year.
*   **Dropout:**
    *   Control: 10% annual dropout rate.
    *   Experimental: 5% annual dropout rate.
*   **Maximum Follow-up:** Each patient is followed for a maximum of 2 years from randomization.

### Defining Input Parameters

```{r}
# Enrollment: rate of 4 patients/month for 5 months -> ~20 patients
enroll_rate <- data.frame(
  rate = 4,
  duration = 5
)

# Failure rates (events per unit time)
fail_rate <- data.frame(
  treatment = c("Control", "Experimental"),
  rate = c(0.5, 0.3) # events per year (assuming time unit is year, adjust enroll duration if needed)
)

# Let's ensure time units are consistent.
# If fail_rate is per year, then durations should be in years.
# 5 months = 5/12 years.
enroll_rate <- data.frame(
  rate = 20 / (5 / 12), # 20 patients over 5/12 years
  duration = 5 / 12
)

# Dropout rates (per year)
dropout_rate <- data.frame(
  treatment = c("Control", "Experimental"),
  rate = c(0.1, 0.05),
  duration = c(100, 100) # constant rate for long duration
)

# Maximum follow-up per patient (years)
max_followup <- 2
```

## Running the Simulation

```{r}
set.seed(123)

sim_data <- nb_sim(
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  dropout_rate = dropout_rate,
  max_followup = max_followup,
  n = 20
)

head(sim_data)
```

The output contains multiple rows per subject:
*   `event = 1`: An actual recurrent event.
*   `event = 0`: The censoring time (either due to dropout or reaching `max_followup`).

## Analyzing the Data

We can aggregate the data to calculate the observed event rates and total follow-up time for each group.

```{r}
# Summarize by treatment group
summary_stats <- sim_data %>%
  group_by(treatment) %>%
  summarise(
    n_subjects = n_distinct(id),
    total_events = sum(event == 1),
    total_followup = sum(tte[event == 0]), # Censoring time represents total individual follow-up
    observed_rate = total_events / total_followup
  )

print(summary_stats)
```

### Plotting Events

We can visualize the events and censoring times for each subject.

```{r, fig.width=7, fig.height=5}
library(ggplot2)

# Create a plot of patient timelines
# Each line is a patient. Points are events. 'x' is censoring.

ggplot(sim_data, aes(x = tte, y = factor(id), color = treatment)) +
  geom_line(aes(group = id), color = "gray80") +
  geom_point(data = subset(sim_data, event == 1), shape = 19, size = 2) +
  geom_point(data = subset(sim_data, event == 0), shape = 4, size = 3) +
  labs(
    title = "Patient Timelines",
    x = "Time from Randomization (Years)",
    y = "Patient ID",
    caption = "Dots = Events, X = Censoring/Dropout"
  ) +
  theme_minimal()
```


