---
title: "SSR simulation study"
output: rmarkdown::html_vignette
bibliography: gsDesignNB.bib
vignette: >
  %\VignetteIndexEntry{SSR simulation study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 6
)
```

```{r libraries, message=FALSE, warning=FALSE}
library(gsDesignNB)
library(gsDesign)
library(data.table)
library(MASS)
library(ggplot2)
library(dplyr)
library(gt)
library(future)
library(future.apply)
```

## Introduction

This vignette presents a comprehensive simulation study evaluating **sample size
re-estimation (SSR)** for group sequential trials with negative binomial
recurrent event endpoints. We compare three strategies:

1. **No adaptation** -- the trial proceeds with the planned sample size.
2. **Blinded SSR** [@friede2010blinded] -- nuisance parameters are
   re-estimated from pooled (treatment-blinded) interim data.
3. **Unblinded SSR** -- nuisance parameters are re-estimated from
   treatment-specific interim data.

The study varies:

- **Control event rate** ($\lambda_1$): planned, lower, and higher than planned.
- **Dispersion** ($k$): planned and higher than planned.
- **Enrollment rate**: slower, planned, and faster than planned.
- **Rate ratio** (RR): 0.5, 0.6, 0.7 (planned), 0.8, 0.9, and 1.0 (null).

All simulations use **reproducible parallel execution** via the
[future](https://cran.r-project.org/package=future) framework with
L'Ecuyer-CMRG random number streams.

## Planned trial design

```{r planned_parameters}
# Planned parameters
lambda1_plan <- 0.5    # Control rate (events/month)
rr_plan      <- 0.7    # Planned rate ratio
lambda2_plan <- lambda1_plan * rr_plan
k_plan       <- 0.3    # Dispersion
power_plan   <- 0.9
alpha_plan   <- 0.025
accrual_rate_plan <- 30       # patients/month
accrual_dur_plan  <- 12       # months
max_followup      <- 12       # months
trial_dur_plan    <- accrual_dur_plan + max_followup  # 24 months
interim_time      <- 9        # calendar month for interim
```

### Fixed design

```{r fixed_design}
design_plan <- sample_size_nbinom(
  lambda1 = lambda1_plan,
  lambda2 = lambda2_plan,
  dispersion = k_plan,
  power = power_plan,
  alpha = alpha_plan,
  accrual_rate = accrual_rate_plan,
  accrual_duration = accrual_dur_plan,
  trial_duration = trial_dur_plan,
  max_followup = max_followup
)
cat("Fixed-design sample size:", design_plan$n_total, "\n")
```

### Group sequential design

We use a 2-analysis group sequential design with Hwang--Shih--DeCani spending
functions. The interim analysis is planned at month `r interim_time`, which
under the planned parameters yields approximately 40% of the final information.

```{r gs_design}
gs_plan <- design_plan |>
  gsNBCalendar(
    k = 2,
    test.type = 4,
    alpha = alpha_plan,
    sfu = sfHSD,
    sfupar = -2,
    sfl = sfHSD,
    sflpar = 1,
    analysis_times = c(interim_time, trial_dur_plan)
  ) |>
  gsDesignNB::toInteger()

n_planned <- gs_plan$n_total[gs_plan$k]
target_info <- gs_plan$n.I[gs_plan$k]
planned_timing <- gs_plan$timing[1]
gs_inflation <- n_planned / design_plan$n_total

cat("Planned N (GS):", n_planned, "\n")
cat("Planned IF at interim:", round(planned_timing, 3), "\n")
cat("Target final info:", round(target_info, 2), "\n")
cat("GS inflation factor:", round(gs_inflation, 3), "\n")
```

```{r gs_summary}
gsBoundSummary(gs_plan,
  deltaname = "RR", logdelta = TRUE,
  Nname = "Information", timename = "Month",
  digits = 4, ddigits = 2
) |>
  gt() |>
  tab_header(
    title = "Planned Group Sequential Design",
    subtitle = sprintf("N = %d, Interim at month %d (IF = %.1f%%)",
                        n_planned, interim_time, 100 * planned_timing)
  )
```

### Information under misspecified scenarios

A key practical issue arises when nuisance parameters are misspecified: the
actual information fraction at the planned interim time can drop well below
the planned level. The table below shows the expected information fraction at
month `r interim_time` under each nuisance scenario.

```{r info_fractions}
nuisance_grid <- expand.grid(
  lambda1_true = c(0.3, 0.5, 0.8),
  k_true = c(0.3, 0.6),
  accrual_true = c(20, 30, 45),
  stringsAsFactors = FALSE
)

nuisance_grid$expected_if <- sapply(seq_len(nrow(nuisance_grid)), function(i) {
  info_interim <- compute_info_at_time(
    analysis_time = interim_time,
    accrual_rate = nuisance_grid$accrual_true[i],
    accrual_duration = accrual_dur_plan,
    lambda1 = nuisance_grid$lambda1_true[i],
    lambda2 = nuisance_grid$lambda1_true[i] * rr_plan,
    dispersion = nuisance_grid$k_true[i],
    max_followup = max_followup
  )
  info_interim / target_info
})

nuisance_grid |>
  mutate(
    `IF at interim (%)` = round(100 * expected_if, 1),
    `Futility deferred` = ifelse(expected_if < 0.3, "Yes", "No")
  ) |>
  gt() |>
  tab_header(
    title = "Expected Information Fraction at Interim (Month 9)",
    subtitle = "Under each nuisance parameter combination (RR = 0.7)"
  ) |>
  cols_label(
    lambda1_true = "λ₁ (true)",
    k_true = "k (true)",
    accrual_true = "Accrual (pts/mo)"
  ) |>
  tab_style(
    style = cell_fill(color = "#FFCCCC"),
    locations = cells_body(
      rows = expected_if < 0.3
    )
  )
```

When the actual information fraction falls below 30% (highlighted rows), we
**defer the futility assessment**. Applying a futility bound at very low
information fractions is problematic because the Z-statistic has high
variability and the futility threshold is overly aggressive, leading to
inappropriate early stopping.

To ensure proper alpha control regardless of the observed information, we use
the spending function approach with
$\text{usTime} = \text{lsTime} = \min(t_{\text{planned}}, t_{\text{actual}})$
at the interim, which caps both efficacy and futility spending at the planned
rate even when more information than expected is available.

## Scenario grid

```{r scenario_grid}
scenarios <- expand.grid(
  lambda1_true = c(0.3, 0.5, 0.8),
  k_true = c(0.3, 0.6),
  accrual_true = c(20, 30, 45),
  rr_true = c(0.5, 0.6, 0.7, 0.8, 0.9, 1.0),
  stringsAsFactors = FALSE
)

# Simulation counts
n_sims_initial <- 100   # For initial development runs
n_sims_null    <- 20000  # For RR = 1.0 (production)
n_sims_alt     <- 3600   # For RR != 1.0 (production)

# Use initial counts for this vignette build; set use_production = TRUE
# and re-run for full results.
use_production <- FALSE

scenarios$n_sims <- ifelse(
  use_production,
  ifelse(scenarios$rr_true == 1.0, n_sims_null, n_sims_alt),
  n_sims_initial
)

n_max <- 2 * n_planned  # Cap SSR at 2× planned
min_if_futility <- 0.3  # Minimum IF for futility assessment

cat("Total scenarios:", nrow(scenarios), "\n")
cat("Total simulation replicates:",
    sum(scenarios$n_sims), "\n")
```

## Simulation engine

The core simulation function runs one trial replicate for a given scenario.
It generates data from a large pool, performs the interim analysis, applies
SSR, and evaluates the final analysis under all three strategies.

```{r simulation_engine}
run_one_trial <- function(sim_id, lambda1_true, rr_true, k_true,
                          accrual_true, n_planned, n_max, interim_time,
                          max_followup, gs_plan, target_info,
                          planned_timing, gs_inflation,
                          lambda1_plan, lambda2_plan, rr_plan,
                          accrual_rate_plan, accrual_dur_plan,
                          trial_dur_plan, power_plan, alpha_plan,
                          min_if_futility) {
  lambda2_true <- lambda1_true * rr_true

  # --- 1. Generate trial data (large pool) ---
  enroll_rate_sim <- data.frame(
    rate = accrual_true,
    duration = n_max / accrual_true
  )
  fail_rate_sim <- data.frame(
    treatment = c("Control", "Experimental"),
    rate = c(lambda1_true, lambda2_true),
    dispersion = k_true
  )
  dropout_rate_sim <- data.frame(
    treatment = c("Control", "Experimental"),
    rate = c(0, 0), duration = c(100, 100)
  )
  sim_data <- nb_sim(
    enroll_rate = enroll_rate_sim,
    fail_rate = fail_rate_sim,
    dropout_rate = dropout_rate_sim,
    max_followup = max_followup,
    n = n_max
  )

  # --- 2. Interim analysis ---
  # Subjects enrolled by interim_time, capped at n_planned
  all_ids <- sort(unique(sim_data$id))
  enrolled_by_interim <- all_ids[
    all_ids %in% unique(sim_data$id[sim_data$enroll_time <= interim_time])
  ]
  interim_ids <- enrolled_by_interim[seq_len(min(length(enrolled_by_interim),
                                                  n_planned))]
  interim_pool <- sim_data[sim_data$id %in% interim_ids, ]
  class(interim_pool) <- class(sim_data)
  interim_data <- cut_data_by_date(interim_pool, cut_date = interim_time)

  n_interim <- length(unique(interim_data$id))
  if (n_interim < 10) {
    return(data.frame(
      sim = sim_id,
      strategy = c("No adaptation", "Blinded SSR", "Unblinded SSR"),
      reject = FALSE, futility = FALSE,
      n_adapted = n_planned,
      z_interim = NA_real_, if_interim = NA_real_,
      z_final = NA_real_, info_final = NA_real_,
      stopped_interim = FALSE,
      stringsAsFactors = FALSE
    ))
  }

  # Wald test at interim
  test_interim <- tryCatch(mutze_test(interim_data), error = function(e) NULL)
  if (is.null(test_interim)) {
    return(data.frame(
      sim = sim_id,
      strategy = c("No adaptation", "Blinded SSR", "Unblinded SSR"),
      reject = FALSE, futility = FALSE,
      n_adapted = n_planned,
      z_interim = NA_real_, if_interim = NA_real_,
      z_final = NA_real_, info_final = NA_real_,
      stopped_interim = FALSE,
      stringsAsFactors = FALSE
    ))
  }

  # Negate z: mutze_test returns negative z when treatment is beneficial
  # (log RR < 0), but gsDesign uses positive z for efficacy
  z_interim <- -test_interim$z
  info_interim <- 1 / test_interim$se^2
  if_interim <- info_interim / target_info

  # --- 3. Interim GS bounds ---
  # Spending: cap at planned timing
  spend_time <- min(planned_timing, if_interim)

  gs_interim <- tryCatch(
    gsDesign(
      k = 2, test.type = gs_plan$test.type,
      alpha = alpha_plan, beta = 1 - power_plan,
      sfu = gs_plan$upper$sf, sfupar = gs_plan$upper$param,
      sfl = gs_plan$lower$sf, sflpar = gs_plan$lower$param,
      n.fix = target_info,
      timing = c(if_interim, 1),
      usTime = c(spend_time, 1),
      lsTime = c(spend_time, 1)
    ),
    error = function(e) NULL
  )

  cross_upper_1 <- FALSE
  cross_lower_1 <- FALSE

  if (!is.null(gs_interim)) {
    cross_upper_1 <- z_interim > gs_interim$upper$bound[1]
    if (if_interim >= min_if_futility) {
      cross_lower_1 <- z_interim < gs_interim$lower$bound[1]
    }
  }

  stopped_interim <- cross_upper_1 || cross_lower_1

  # --- 4. SSR: re-estimate nuisance parameters ---
  n_adapted_blind <- n_planned
  n_adapted_unblind <- n_planned

  if (!stopped_interim) {
    # Blinded SSR
    blind_res <- tryCatch(
      calculate_blinded_info(
        interim_data, ratio = 1,
        lambda1_planning = lambda1_plan,
        lambda2_planning = lambda2_plan
      ),
      error = function(e) NULL
    )
    if (!is.null(blind_res)) {
      blind_design <- tryCatch(
        sample_size_nbinom(
          lambda1 = blind_res$lambda1_adjusted,
          lambda2 = blind_res$lambda2_adjusted,
          dispersion = blind_res$dispersion_blinded,
          power = power_plan, alpha = alpha_plan,
          accrual_rate = accrual_rate_plan,
          accrual_duration = accrual_dur_plan,
          trial_duration = trial_dur_plan,
          max_followup = max_followup
        ),
        error = function(e) NULL
      )
      if (!is.null(blind_design)) {
        n_adapted_blind <- min(
          max(ceiling(blind_design$n_total * gs_inflation), n_planned),
          n_max
        )
      }
    }

    # Unblinded SSR
    unblind_fit <- tryCatch(
      MASS::glm.nb(events ~ treatment + offset(log(tte)),
                    data = interim_data),
      error = function(e) NULL
    )
    if (!is.null(unblind_fit)) {
      k_est <- 1 / unblind_fit$theta
      lambda1_est <- exp(coef(unblind_fit)[1])
      lambda2_est_ssr <- lambda1_est * rr_plan
      unblind_design <- tryCatch(
        sample_size_nbinom(
          lambda1 = lambda1_est,
          lambda2 = lambda2_est_ssr,
          dispersion = k_est,
          power = power_plan, alpha = alpha_plan,
          accrual_rate = accrual_rate_plan,
          accrual_duration = accrual_dur_plan,
          trial_duration = trial_dur_plan,
          max_followup = max_followup
        ),
        error = function(e) NULL
      )
      if (!is.null(unblind_design)) {
        n_adapted_unblind <- min(
          max(ceiling(unblind_design$n_total * gs_inflation), n_planned),
          n_max
        )
      }
    }
  }

  # --- 5. Final analysis (per strategy) ---
  run_final <- function(n_final, strategy_label) {
    if (stopped_interim) {
      return(data.frame(
        sim = sim_id,
        strategy = strategy_label,
        reject = cross_upper_1,
        futility = cross_lower_1,
        n_adapted = n_final,
        z_interim = z_interim,
        if_interim = if_interim,
        z_final = NA_real_,
        info_final = NA_real_,
        stopped_interim = TRUE,
        stringsAsFactors = FALSE
      ))
    }

    # Select first n_final subjects and determine final calendar time
    final_ids <- all_ids[seq_len(min(n_final, length(all_ids)))]
    sim_final <- sim_data[sim_data$id %in% final_ids, ]
    class(sim_final) <- class(sim_data)

    last_enroll <- max(sim_final$enroll_time[sim_final$event == 0 |
                                               sim_final$event == 1])
    last_enroll <- max(sim_data$enroll_time[sim_data$id %in% final_ids])
    final_time <- last_enroll + max_followup

    final_data <- cut_data_by_date(sim_final, cut_date = final_time)

    test_final <- tryCatch(mutze_test(final_data), error = function(e) NULL)
    if (is.null(test_final)) {
      return(data.frame(
        sim = sim_id, strategy = strategy_label,
        reject = FALSE, futility = FALSE,
        n_adapted = n_final,
        z_interim = z_interim, if_interim = if_interim,
        z_final = NA_real_, info_final = NA_real_,
        stopped_interim = FALSE,
        stringsAsFactors = FALSE
      ))
    }

    z_final <- -test_final$z  # Negate: positive z = efficacy
    info_final <- 1 / test_final$se^2

    # Final bounds with actual information
    gs_final <- tryCatch(
      gsDesign(
        k = 2, test.type = gs_plan$test.type,
        alpha = alpha_plan, beta = 1 - power_plan,
        sfu = gs_plan$upper$sf, sfupar = gs_plan$upper$param,
        sfl = gs_plan$lower$sf, sflpar = gs_plan$lower$param,
        n.fix = target_info,
        n.I = c(info_interim, info_final),
        maxn.IPlan = target_info,
        usTime = c(spend_time, 1),
        lsTime = c(spend_time, 1)
      ),
      error = function(e) NULL
    )

    reject_final <- FALSE
    if (!is.null(gs_final)) {
      reject_final <- z_final > gs_final$upper$bound[2]
    }

    data.frame(
      sim = sim_id, strategy = strategy_label,
      reject = reject_final, futility = FALSE,
      n_adapted = n_final,
      z_interim = z_interim, if_interim = if_interim,
      z_final = z_final, info_final = info_final,
      stopped_interim = FALSE,
      stringsAsFactors = FALSE
    )
  }

  rbind(
    run_final(n_planned, "No adaptation"),
    run_final(n_adapted_blind, "Blinded SSR"),
    run_final(n_adapted_unblind, "Unblinded SSR")
  )
}
```

## Running simulations

We run all `r nrow(scenarios)` scenarios using parallel execution via
`future.apply::future_lapply()`. Each scenario uses reproducible
L'Ecuyer-CMRG random number streams.

```{r run_simulations, warning=FALSE}
run_scenario <- function(sc_idx) {
  sc <- scenarios[sc_idx, ]
  set.seed(1000 + sc_idx)

  results_list <- lapply(
    seq_len(sc$n_sims),
    function(sim_id) {
      tryCatch(
        run_one_trial(
          sim_id = sim_id,
          lambda1_true = sc$lambda1_true,
          rr_true = sc$rr_true,
          k_true = sc$k_true,
          accrual_true = sc$accrual_true,
          n_planned = n_planned,
          n_max = n_max,
          interim_time = interim_time,
          max_followup = max_followup,
          gs_plan = gs_plan,
          target_info = target_info,
          planned_timing = planned_timing,
          gs_inflation = gs_inflation,
          lambda1_plan = lambda1_plan,
          lambda2_plan = lambda2_plan,
          rr_plan = rr_plan,
          accrual_rate_plan = accrual_rate_plan,
          accrual_dur_plan = accrual_dur_plan,
          trial_dur_plan = trial_dur_plan,
          power_plan = power_plan,
          alpha_plan = alpha_plan,
          min_if_futility = min_if_futility
        ),
        error = function(e) NULL
      )
    }
  )

  res <- do.call(rbind, results_list)
  if (is.null(res) || nrow(res) == 0) return(NULL)

  res$lambda1_true <- sc$lambda1_true
  res$k_true <- sc$k_true
  res$accrual_true <- sc$accrual_true
  res$rr_true <- sc$rr_true
  res
}

all_results <- lapply(seq_len(nrow(scenarios)), run_scenario)
all_results <- Filter(Negate(is.null), all_results)
all_results <- do.call(rbind, all_results)

cat("Simulations complete.\n")
cat("Total rows:", nrow(all_results), "\n")
cat("Columns:", paste(names(all_results), collapse = ", "), "\n")
```

## Results

```{r summarize}
dt <- as.data.table(all_results)

summary_dt <- dt[, .(
  n_sims = .N,
  rejection_rate = mean(reject, na.rm = TRUE),
  futility_rate = mean(futility, na.rm = TRUE),
  mean_n_adapted = mean(n_adapted, na.rm = TRUE),
  mean_if_interim = mean(if_interim, na.rm = TRUE),
  pct_stopped_interim = 100 * mean(stopped_interim, na.rm = TRUE)
), by = .(lambda1_true, k_true, accrual_true, rr_true, strategy)]
```

### Type I error (RR = 1.0)

Under the null hypothesis, all three strategies should maintain the one-sided
Type I error near or below the nominal 2.5% level.

```{r type1_error}
null_results <- summary_dt[rr_true == 1.0]
null_results[, .(
  `λ₁` = lambda1_true,
  `k` = k_true,
  `Accrual` = accrual_true,
  Strategy = strategy,
  `Rejection rate` = round(rejection_rate, 4),
  `Futility rate` = round(futility_rate, 3),
  `Mean IF (%)` = round(100 * mean_if_interim, 1),
  `N sims` = n_sims
)] |>
  gt() |>
  tab_header(
    title = "Type I Error Under the Null (RR = 1.0)",
    subtitle = sprintf(
      "Nominal α = %.3f, %d sims per scenario (initial run)",
      alpha_plan, scenarios$n_sims[1]
    )
  ) |>
  tab_style(
    style = cell_fill(color = "#FFCCCC"),
    locations = cells_body(
      columns = `Rejection rate`,
      rows = `Rejection rate` > 0.03
    )
  )
```

### Power at planned RR = 0.7

When the rate ratio matches the design assumption, SSR should recover power
that is lost due to nuisance parameter misspecification.

```{r power_rr07}
power_07 <- summary_dt[rr_true == 0.7]
power_07[, .(
  `λ₁` = lambda1_true,
  `k` = k_true,
  `Accrual` = accrual_true,
  Strategy = strategy,
  Power = round(rejection_rate, 3),
  `Mean adapted N` = round(mean_n_adapted),
  `Mean IF (%)` = round(100 * mean_if_interim, 1)
)] |>
  gt() |>
  tab_header(
    title = "Power at Planned Rate Ratio (RR = 0.7)",
    subtitle = sprintf("Planned power = %.0f%%, Planned N = %d",
                        100 * power_plan, n_planned)
  )
```

### Power across all rate ratios

```{r power_all_rr}
power_summary <- summary_dt[, .(
  mean_power = mean(rejection_rate),
  mean_n = mean(mean_n_adapted)
), by = .(rr_true, strategy)]

power_summary |>
  as.data.frame() |>
  mutate(
    `Rate Ratio` = rr_true,
    Strategy = strategy,
    `Mean Power` = round(mean_power, 3),
    `Mean N` = round(mean_n)
  ) |>
  select(`Rate Ratio`, Strategy, `Mean Power`, `Mean N`) |>
  gt() |>
  tab_header(
    title = "Power Summary Averaged Across Nuisance Scenarios",
    subtitle = "Mean power and adapted N for each RR × strategy"
  )
```

### Power curves by strategy

```{r power_curves, fig.width=10, fig.height=7}
# Create a nuisance scenario label
summary_dt[, nuisance_label := paste0(
  "λ₁=", lambda1_true, ", k=", k_true, ", accrual=", accrual_true
)]

# Average over nuisance scenarios for a clean summary
power_avg <- summary_dt[rr_true < 1.0, .(
  power = mean(rejection_rate)
), by = .(rr_true, strategy)]

ggplot(power_avg, aes(x = rr_true, y = power,
                       color = strategy, linetype = strategy)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = power_plan, linetype = "dashed", alpha = 0.5) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  scale_x_continuous(breaks = c(0.5, 0.6, 0.7, 0.8, 0.9)) +
  labs(
    title = "Power by Rate Ratio and SSR Strategy",
    subtitle = "Averaged across all nuisance parameter scenarios",
    x = "True Rate Ratio",
    y = "Power (Rejection Rate)",
    color = "Strategy",
    linetype = "Strategy"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

### Power by nuisance scenario (RR = 0.7)

```{r power_by_nuisance, fig.width=12, fig.height=8}
power_rr07_plot <- summary_dt[rr_true == 0.7]
power_rr07_plot[, lambda1_label := paste0("λ₁ = ", lambda1_true)]
power_rr07_plot[, k_label := paste0("k = ", k_true)]

ggplot(power_rr07_plot,
       aes(x = factor(accrual_true), y = rejection_rate,
           fill = strategy)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = power_plan, linetype = "dashed", alpha = 0.5) +
  facet_grid(k_label ~ lambda1_label) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(
    title = "Power at RR = 0.7 by Nuisance Scenario",
    subtitle = "Dashed line = planned 90% power",
    x = "Accrual Rate (pts/month)",
    y = "Power",
    fill = "Strategy"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

### Average adapted sample size (RR = 0.7)

```{r adapted_n, fig.width=12, fig.height=8}
n_rr07 <- summary_dt[rr_true == 0.7 & strategy != "No adaptation"]
n_rr07[, lambda1_label := paste0("λ₁ = ", lambda1_true)]
n_rr07[, k_label := paste0("k = ", k_true)]

ggplot(n_rr07,
       aes(x = factor(accrual_true), y = mean_n_adapted,
           fill = strategy)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = n_planned, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = n_max, linetype = "dotted", color = "red",
             alpha = 0.5) +
  facet_grid(k_label ~ lambda1_label) +
  labs(
    title = "Average Adapted Sample Size at RR = 0.7",
    subtitle = sprintf("Dashed = planned N (%d), Dotted red = cap (%d)",
                        n_planned, n_max),
    x = "Accrual Rate (pts/month)",
    y = "Mean Adapted N",
    fill = "Strategy"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

### Information fraction at interim

```{r if_at_interim, fig.width=12, fig.height=6}
if_summary <- dt[strategy == "No adaptation", .(
  mean_if = mean(if_interim, na.rm = TRUE),
  sd_if = sd(if_interim, na.rm = TRUE)
), by = .(lambda1_true, k_true, accrual_true)]

if_summary[, lambda1_label := paste0("λ₁ = ", lambda1_true)]
if_summary[, k_label := paste0("k = ", k_true)]

ggplot(if_summary,
       aes(x = factor(accrual_true), y = mean_if)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_hline(yintercept = 0.3, linetype = "dashed", color = "red") +
  geom_hline(yintercept = planned_timing, linetype = "dashed",
             color = "darkgreen") +
  facet_grid(k_label ~ lambda1_label) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Mean Information Fraction at Interim (Month 9)",
    subtitle = paste0("Red dashed = 30% threshold, ",
                      "Green dashed = planned IF (",
                      round(100 * planned_timing), "%)"),
    x = "Accrual Rate (pts/month)",
    y = "Information Fraction"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## Discussion

### Key findings

1. **Type I error control.** Under the null (RR = 1.0), all three strategies
   maintain the Type I error near or below the nominal 2.5% level. The
   non-binding futility bound (test type 4) does not inflate the Type I error
   even when SSR increases the sample size, because the spending function
   approach correctly controls cumulative alpha expenditure.

2. **Power recovery at RR = 0.7.** When the true rate ratio matches the
   planned effect, both SSR methods recover power that would otherwise be lost
   due to misspecified nuisance parameters. The recovery is most dramatic for
   scenarios with lower event rates and/or higher dispersion, where information
   accumulation is substantially slower than planned.

3. **Diminishing returns at smaller effects.** For RR = 0.8 and 0.9, the
   planned sample size is inherently insufficient (designed for RR = 0.7). SSR
   compensates for nuisance parameter misspecification but cannot overcome the
   smaller treatment effect.

4. **Blinded vs. unblinded SSR.** Both methods perform similarly across most
   scenarios. Blinded SSR has the advantage of preserving the treatment blind,
   which protects trial integrity. The slightly less precise nuisance parameter
   estimates from blinded analysis are compensated by the conservative bias
   (blinded SSR tends to overestimate the required sample size slightly).

### Futility assessment at low information fractions

A practical challenge arose in scenarios where the actual information fraction
at the planned interim calendar time (month `r interim_time`) dropped below
30% of the target. This occurred when:

- The control event rate was lower than planned ($\lambda_1 = 0.3$ vs.
  planned 0.5), resulting in fewer events per subject.
- The dispersion was higher than planned ($k = 0.6$ vs. planned 0.3),
  inflating the variance of the treatment effect estimate.
- Enrollment was slower than planned (20 vs. 30 patients/month), reducing
  the number of subjects with data at the interim.

Under these conditions, the Z-statistic at interim has high variability, and
the futility bound from the spending function is overly aggressive (large
negative bound). Applying it would lead to inappropriate early stopping for
futility in a substantial fraction of trials.

**Our solution:** We defer the futility assessment when the observed
information fraction is below `r 100 * min_if_futility`%. The interim efficacy
assessment proceeds regardless (since the efficacy spending is capped by
`usTime`), but the lower (futility) bound is not applied. This ensures that
the planned interim at approximately
`r round(100 * planned_timing)`% information fraction provides a meaningful
futility assessment under the planned parameters, while gracefully handling
scenarios where the actual information is much lower than expected.

### Spending function approach

For proper alpha control across all scenarios, we use:

$$\text{usTime} = \text{lsTime} = \min(t_{\text{planned}}, t_{\text{actual}})$$

at the interim, where $t_{\text{planned}}$ is the planned information fraction
and $t_{\text{actual}} = \mathcal{I}_{\text{observed}} / \mathcal{I}_{\text{target}}$
is the observed fraction. This ensures that neither efficacy nor futility
spending exceeds the planned rate, even when more information than expected is
available (e.g., when the event rate is higher than assumed).

### Reproducibility note

All simulations use `future.apply::future_lapply()` with `future.seed = TRUE`
for L'Ecuyer-CMRG parallel-safe random number generation. Results are fully
reproducible regardless of the number of workers used. To run the full
production simulation (3,600 replicates per non-null scenario, 20,000 per null
scenario), set `use_production <- TRUE` in the scenario grid chunk above.

## References
