---
title: "SSR simulation study"
output: rmarkdown::html_vignette
bibliography: gsDesignNB.bib
vignette: >
  %\VignetteIndexEntry{SSR simulation study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 7
)
design_note <- "Design summary computed below"
```

```{r libraries, message=FALSE, warning=FALSE}
library(gsDesignNB)
library(gsDesign)
library(data.table)
library(MASS)
library(ggplot2)
library(dplyr)
library(gt)
library(DT)
library(future)
library(future.apply)
```

## Introduction

This vignette presents a simulation study evaluating **sample size
re-estimation (SSR)** for group sequential trials with negative binomial
recurrent event endpoints. We compare three strategies:

1. **No adaptation** -- the trial proceeds with the planned sample size.
2. **Blinded SSR** [@friede2010blinded] -- nuisance parameters are
   re-estimated from pooled (treatment-blinded) interim data.
3. **Unblinded SSR** -- nuisance parameters are re-estimated from
   treatment-specific interim data.

For hypothesis testing, all interim and final analyses use the
`mutze_test()` Wald statistic. This test underpins both efficacy/futility
crossing decisions and the unblinded-information fallback used when blinded
information estimation fails.

*Note:* Much of the simulation code in this vignette was generated with the
assistance of a large language model (LLM) and then reviewed and refined
by the package authors.

Interim analyses are triggered when the **blinded statistical information**
reaches the planned fraction of the target, rather than at fixed calendar
times. This ensures analyses occur at comparable information levels
regardless of the true nuisance parameters.

## Planned trial design

```{r planned_parameters}
lambda1_plan      <- 0.5
rr_plan           <- 0.7
lambda2_plan      <- lambda1_plan * rr_plan
k_plan            <- 0.5
power_plan        <- 0.9
alpha_plan        <- 0.025
accrual_rate_plan <- 30
accrual_scenario_plan <- 18
accrual_dur_plan  <- 12
max_followup      <- 12
trial_dur_plan    <- accrual_dur_plan + max_followup
event_gap_val     <- 20 / 30.4375  # 20-day gap between events
analysis_times_plan <- c(9, 14, 24)  # Calendar times under plan
```

### Fixed design

```{r fixed_design}
design_plan <- sample_size_nbinom(
  lambda1 = lambda1_plan, lambda2 = lambda2_plan,
  dispersion = k_plan, power = power_plan, alpha = alpha_plan,
  accrual_rate = accrual_rate_plan,
  accrual_duration = accrual_dur_plan,
  trial_duration = trial_dur_plan,
  max_followup = max_followup,
  event_gap = event_gap_val
)
cat("Fixed-design sample size:", design_plan$n_total, "\n")
```

### Group sequential design

We use a 3-analysis group sequential design with HSD efficacy spending and
Cauchy futility spending calibrated so the futility bounds correspond to
approximately RR = 0.9 at each analysis.

Specifically, `sfl = sfCauchy` with `sflpar = c(0.4, 0.75, 0.56, 0.63)` gives
planned lower bounds that are close to declaring futility when the observed RR
is greater than about 0.9 at either interim. Because the design uses
non-binding futility (`test.type = 4`), ignoring a lower-bound crossing does
not invalidate the final efficacy decision, which is still based on upper
boundary control.

```{r gs_design}
gs_plan <- design_plan |>
  gsNBCalendar(
    k = 3, test.type = 4, alpha = alpha_plan,
    sfu = sfHSD, sfupar = -2,
    sfl = sfCauchy, sflpar = c(0.4, 0.75, 0.56, 0.63),
    analysis_times = analysis_times_plan
  ) |>
  gsDesignNB::toInteger()

n_planned <- gs_plan$n_total[gs_plan$k]
target_info <- gs_plan$n.I[gs_plan$k]
planned_timing <- gs_plan$timing
gs_inflation <- n_planned / design_plan$n_total
accrual_rate_plan_eff <- n_planned / accrual_dur_plan
design_note <- paste0(
  "Design: lambda1=", lambda1_plan,
  ", RR=", rr_plan,
  ", k=", k_plan,
  ", planned accrual=", round(accrual_rate_plan_eff, 1), "/mo",
  ", planned N=", n_planned,
  ", max follow-up=", max_followup, " mo"
)

```

The group sequential inflation factor is
`r round(gs_inflation, 3)`, giving a planned sample size of
`r n_planned` (up from the fixed-design
`r round(design_plan$n_total)`) with an effective accrual rate of
`r round(accrual_rate_plan_eff, 1)` patients per month.

```{r gs_bound_summary}
gsBoundSummary(gs_plan,
  deltaname = "RR", logdelta = TRUE,
  Nname = "Information", timename = "Month",
  digits = 4, ddigits = 2
) |>
  as.data.frame() |>
  gt() |>
  tab_header(
    title = "Planned Group Sequential Design",
    subtitle = design_note
  ) |>
  tab_footnote(
    "Cauchy futility spending gives planned futility near observed RR > 0.9 at IA1/IA2; lower bounds are non-binding."
  ) |>
  tab_footnote(
    footnote = sprintf(
      "Planned cumulative sample size: IA1 = %.0f, IA2 = %.0f, Final = %.0f.",
      gs_plan$n_total[1], gs_plan$n_total[2], gs_plan$n_total[3]
    )
  )
```

### Group sequential sample size under each nuisance scenario

The figure below shows how the required group sequential sample size and
monthly enrollment rate vary with the control rate $\lambda_1$ and
dispersion $k$.
Since the sample size does not depend on the enrollment pace (accrual rate
only scales enrollment duration), we show two lines -- one per $k$ value --
with $\lambda_1$ on the horizontal axis.
The planned design assumptions ($\lambda_1 = `r lambda1_plan`$,
$k = `r k_plan`$) are marked with a dashed line.

```{r gs_n_by_scenario, fig.width=9, fig.height=4.5}
lambda1_seq <- seq(0.2, 1.0, by = 0.1)
gs_n_grid <- expand.grid(
  lambda1_true = lambda1_seq,
  k_true = c(0.5, 1.0),
  stringsAsFactors = FALSE
)

gs_n_grid$GS_N <- sapply(seq_len(nrow(gs_n_grid)), function(i) {
  fixed_i <- tryCatch(
    sample_size_nbinom(
      lambda1 = gs_n_grid$lambda1_true[i],
      lambda2 = gs_n_grid$lambda1_true[i] * rr_plan,
      dispersion = gs_n_grid$k_true[i],
      power = power_plan, alpha = alpha_plan,
      accrual_rate = accrual_rate_plan,
      accrual_duration = accrual_dur_plan,
      trial_duration = trial_dur_plan,
      max_followup = max_followup,
      event_gap = event_gap_val
    ),
    error = function(e) NULL
  )
  if (is.null(fixed_i)) return(NA_real_)
  tryCatch({
    g <- gsNBCalendar(fixed_i, k = 3, test.type = 4, alpha = alpha_plan,
      sfu = sfHSD, sfupar = -2,
      sfl = sfCauchy, sflpar = c(0.4, 0.75, 0.56, 0.63),
      analysis_times = analysis_times_plan) |>
      gsDesignNB::toInteger()
    g$n_total[g$k]
  }, error = function(e) NA_real_)
})

gs_n_grid$Accrual_rate <- gs_n_grid$GS_N / accrual_dur_plan
gs_n_grid$k_label <- paste0("k = ", gs_n_grid$k_true)
gs_n_grid$N_label <- round(gs_n_grid$GS_N)
gs_n_grid$Rate_label <- round(gs_n_grid$Accrual_rate)

plot_n <- ggplot(gs_n_grid, aes(x = lambda1_true, y = GS_N,
                                color = k_label, linetype = k_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_text(aes(label = N_label), vjust = -0.8, size = 3.2, show.legend = FALSE) +
  geom_vline(xintercept = lambda1_plan, linetype = "dashed", alpha = 0.4) +
  labs(x = expression(lambda[1]), y = "GS sample size (N)",
       color = NULL, linetype = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")

plot_rate <- ggplot(gs_n_grid, aes(x = lambda1_true, y = Accrual_rate,
                                   color = k_label, linetype = k_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_text(aes(label = Rate_label), vjust = -0.8, size = 3.2, show.legend = FALSE) +
  geom_vline(xintercept = lambda1_plan, linetype = "dashed", alpha = 0.4) +
  labs(x = expression(lambda[1]), y = "Enrollment rate (pts/mo)",
       color = NULL, linetype = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")

gridExtra::grid.arrange(plot_n, plot_rate, ncol = 2,
  top = grid::textGrob(
    "GS Sample Size and Enrollment Rate by Control Rate and Dispersion",
    gp = grid::gpar(fontsize = 14, fontface = "bold")
  )
)
```

Across the range of $\lambda_1$ and $k$ values considered, the required
group sequential sample size ranges from
`r min(gs_n_grid$GS_N, na.rm = TRUE)` to
`r max(gs_n_grid$GS_N, na.rm = TRUE)`.
Higher dispersion ($k = 0.6$) consistently requires more subjects than
$k = 0.3$, while higher control rates reduce the required N because each
subject contributes more information.
This wide range highlights why sample size re-estimation can be valuable
when nuisance parameters are uncertain.

### Expected information fraction at planned time of each interim

This table shows the expected information fraction at each planned interim
calendar time under each nuisance scenario. Under the approximate design assumptions
(bold green row), the information fractions match the GS design reasonably well
(`r paste(round(100 * planned_timing[1:2], 1), collapse = "%, ")`%).

```{r info_fractions}
nuisance_grid <- expand.grid(
  lambda1_true = c(0.3, 0.5, 0.8),
  k_true = c(0.5, 1.0),
  accrual_true = c(12, 18, 24),
  stringsAsFactors = FALSE
)

# Accrual scenarios use effective monthly enrollment rates directly.

for (a in 1:2) {
  col_name <- paste0("IF_analysis_", a)
  nuisance_grid[[col_name]] <- sapply(seq_len(nrow(nuisance_grid)), function(i) {
    accrual_eff <- nuisance_grid$accrual_true[i]
    info_at_t <- compute_info_at_time(
      analysis_time = analysis_times_plan[a],
      accrual_rate = accrual_eff,
      accrual_duration = accrual_dur_plan,
      lambda1 = nuisance_grid$lambda1_true[i],
      lambda2 = nuisance_grid$lambda1_true[i] * rr_plan,
      dispersion = nuisance_grid$k_true[i],
      max_followup = max_followup,
      event_gap = event_gap_val
    )
    round(100 * info_at_t / target_info, 1)
  })
}

nuisance_grid |>
  gt() |>
  tab_header(
    title = "Expected Information Fraction (%) at Planned Time of Each Interim",
    subtitle = design_note
  ) |>
  cols_label(
    lambda1_true = "lambda1",
    k_true = "k",
    accrual_true = "Accrual (pts/mo)",
    IF_analysis_1 = paste0("IA 1 (mo ", analysis_times_plan[1], ")"),
    IF_analysis_2 = paste0("IA 2 (mo ", analysis_times_plan[2], ")")
  ) |>
  tab_footnote(
    footnote = paste(
      "Computed via compute_info_at_time() divided by planned final information.",
      "Accrual values (12/18/24) are effective enrollment rates used directly.",
      "Bold green = design assumptions.",
      "With information-based timing, interims occur when blinded info reaches",
      "the target fraction, so the calendar time varies by scenario."
    )
  )
```

## Scenario grid

```{r scenario_grid}
scenarios <- expand.grid(
  lambda1_true = c(0.3, 0.5, 0.8),
  k_true = c(0.5, 1.0),
  accrual_true = c(12, 18, 24),
  rr_true = c(0.6, 0.7, 0.85, 1.0, 1.1),
  stringsAsFactors = FALSE
)

n_sims_initial <- 50
use_production <- identical(Sys.getenv("GSDESIGNNB_PRODUCTION_SSR"), "true")

scenarios$n_sims <- ifelse(
  use_production,
  3600,
  n_sims_initial
)
scenarios$accrual_eff <- scenarios$accrual_true

n_max <- 2 * n_planned
min_if_futility <- 0.3
target_if <- planned_timing  # Target IF for each analysis
# IA2 adaptation gate (less strict than prior 80% / 2 months setting)
max_enrollment_frac_for_ia2 <- 1.00
min_months_to_close_for_adapt <- 2
analysis_lag_months <- 2

# Optional precomputed outputs for fast vignette builds
precomputed_basename <- "ssr_sim_vignette_outputs.rds"
project_root <- if (file.exists("DESCRIPTION")) "." else
  if (file.exists("../DESCRIPTION")) ".." else "."
precomputed_source_path <- file.path(
  project_root, "inst", "extdata", precomputed_basename
)
precomputed_file <- system.file("extdata", precomputed_basename, package = "gsDesignNB")
if (precomputed_file == "" && file.exists(precomputed_source_path)) {
  precomputed_file <- precomputed_source_path
}
force_rerun <- identical(Sys.getenv("GSDESIGNNB_FORCE_SSR_SIM"), "true")
use_precomputed <- (!use_production) && !force_rerun && nzchar(precomputed_file)
save_precomputed <- identical(Sys.getenv("GSDESIGNNB_SAVE_SSR_OUTPUTS"), "true")
save_precomputed_path <- precomputed_source_path

cat("Scenarios:", nrow(scenarios), "| Replicates:", sum(scenarios$n_sims), "\n")
cat("Accrual rates used in simulation:", paste(round(sort(unique(scenarios$accrual_true)), 1), collapse = ", "),
    "/month\n")
cat(
  "IA2 SSR gate: adaptation uses cutoff at min(IA2 time, predicted close - ",
  min_months_to_close_for_adapt,
  " months); enrollment cap <= ",
  100 * max_enrollment_frac_for_ia2,
  "%.\n",
  sep = ""
)
cat("Futility-stop sample size counts enrollment through +",
    analysis_lag_months, " months after stop.\n", sep = "")
cat("Production plan: 3600 reps for each main-grid scenario;\n")
cat("separate RR=1 non-binding futility check uses 20000 reps.\n")
if (use_precomputed) cat("Using precomputed outputs:", precomputed_file, "\n")
```

## Simulation engine

Interim analyses are information-based and use dynamic spending:

1. **IA1 (~40% IF)**: efficacy/futility only; no SSR adaptation.
2. **IA2 (target ~76% IF)**: efficacy/futility evaluated using observed IF
   and spending.
3. **SSR at IA2 only**: nuisance estimates for adaptation use an IA2
   adaptation cutoff time at
   $\min(\text{IA2 time},\; \text{predicted enrollment close} -
   `r min_months_to_close_for_adapt` \text{ months})$, enforcing at least
   `r min_months_to_close_for_adapt` months of operational lead-time.
   The enrollment-fraction cap is set in the scenario setup.

At each analysis, bounds are recalculated from observed information and
spending time. The blinded NB information estimate from
`calculate_blinded_info()` is used to determine interim timing.
If this estimate is unavailable or non-positive, timing falls back to
unblinded information from `mutze_test()` (tracked as fallback counts
in the results table).

```{r simulation_engine}
run_one_trial <- function(sim_id, lambda1_true, rr_true, k_true,
                          accrual_true, accrual_eff, n_planned, n_max,
                          max_followup, gs_plan, target_info,
                          planned_timing, gs_inflation,
                          lambda1_plan, lambda2_plan, rr_plan,
                          accrual_rate_plan, accrual_dur_plan,
                          trial_dur_plan, power_plan, alpha_plan,
                          min_if_futility,
                          max_enrollment_frac_for_ia2,
                          min_months_to_close_for_adapt,
                          analysis_lag_months,
                          event_gap = 0,
                          ignore_futility = FALSE) {
  lambda2_true <- lambda1_true * rr_true
  k_analyses <- 3

  enroll_rate_sim <- data.frame(
    rate = accrual_eff, duration = n_max / accrual_eff
  )
  fail_rate_sim <- data.frame(
    treatment = c("Control", "Experimental"),
    rate = c(lambda1_true, lambda2_true),
    dispersion = k_true
  )
  dropout_rate_sim <- data.frame(
    treatment = c("Control", "Experimental"),
    rate = c(0, 0), duration = c(100, 100)
  )
  sim_data <- nb_sim(
    enroll_rate = enroll_rate_sim, fail_rate = fail_rate_sim,
    dropout_rate = dropout_rate_sim, max_followup = max_followup,
    n = n_max, event_gap = event_gap
  )

  id_enroll <- unique(sim_data[, c("id", "enroll_time")])
  id_enroll <- id_enroll[order(id_enroll$enroll_time, id_enroll$id), ]
  all_ids <- id_enroll$id
  max_calendar <- max(sim_data$calendar_time) + 1

  n_strategies <- c(
    "No adaptation" = n_planned,
    "Blinded SSR" = n_planned,
    "Unblinded SSR" = n_planned
  )

  # Helper: info and method at a given cut time
  info_method_at_t <- function(pool, t) {
    cut_dt <- cut_data_by_date(pool, cut_date = t, event_gap = event_gap)
    if (sum(cut_dt$events) < 2 || length(unique(cut_dt$id)) < 4) {
      return(list(info = 0, method = "insufficient"))
    }

    blind_res <- tryCatch(
      calculate_blinded_info(
        cut_dt, ratio = 1,
        lambda1_planning = lambda1_plan,
        lambda2_planning = lambda2_plan,
        event_gap = event_gap
      ),
      error = function(e) NULL
    )
    if (!is.null(blind_res) &&
        is.finite(blind_res$blinded_info) &&
        !is.na(blind_res$blinded_info) &&
        blind_res$blinded_info > 0) {
      return(list(info = blind_res$blinded_info, method = "blinded"))
    }

    mutze_res <- tryCatch(mutze_test(cut_dt), error = function(e) NULL)
    if (!is.null(mutze_res) &&
        is.finite(mutze_res$se) &&
        !is.na(mutze_res$se) &&
        mutze_res$se > 0) {
      return(list(info = 1 / mutze_res$se^2, method = "unblinded"))
    }

    # User-requested last-resort fallback
    list(info = 100, method = "fallback100")
  }

  # --- Find information-based interim times ---
  # Restrict to planned N for interim timing search
  planned_pool <- sim_data[sim_data$id %in% all_ids[seq_len(n_planned)], ]
  class(planned_pool) <- class(sim_data)

  ia_times <- numeric(k_analyses)
  ia_info_method <- rep("blinded", k_analyses)

  for (a in 1:(k_analyses - 1)) {
    target_info_a <- planned_timing[a] * target_info
    min_t <- if (a == 1) 1 else ia_times[a - 1] + 0.5

    ia_times[a] <- tryCatch(
      get_cut_date(
        data = planned_pool,
        target_info = target_info_a,
        event_gap = event_gap,
        lambda1 = lambda1_plan, lambda2 = lambda2_plan,
        ratio = 1, min_date = min_t, max_date = max_calendar
      ),
      error = function(e) max_calendar
    )

    # Track whether blinded or fallback method was needed at this cut
    ia_info_method[a] <- info_method_at_t(planned_pool, ia_times[a])$method
  }

  # --- Run each strategy ---
  strategy_results <- list()

  for (strat in names(n_strategies)) {
    stopped <- FALSE
    reject <- FALSE
    futility_stop <- FALSE
    reject_stage <- "No reject"
    futility_stage <- "No futility"
    ia2_enroll_frac <- NA_real_
    ia2_months_to_close_pred <- NA_real_
    ia2_adapt_allowed <- FALSE
    ia2_adapt_applied <- FALSE
    ia2_adapt_cut_time <- NA_real_
    n_futility_report <- NA_real_
    z_values <- rep(NA_real_, k_analyses)
    info_values <- rep(NA_real_, k_analyses)
    if_values <- rep(NA_real_, k_analyses)
    cal_times <- rep(NA_real_, k_analyses)

    for (a in seq_len(k_analyses)) {
      if (stopped) break

      current_n <- n_strategies[[strat]]

      if (a < k_analyses) {
        # Interim analyses are based on planned N for all strategies.
        # SSR changes the final sample size target after IA2.
        cut_time <- ia_times[a]
        interim_n <- min(n_planned, length(all_ids))
        analysis_ids <- all_ids[seq_len(interim_n)]
        enrolled_by_t <- id_enroll$id[
          id_enroll$id %in% analysis_ids & id_enroll$enroll_time <= cut_time
        ]
        analysis_pool <- sim_data[sim_data$id %in% enrolled_by_t, ]

        if (a == 2) {
          # Adaptation cutoff may be earlier than IA2 efficacy/futility
          # to ensure time to implement enrollment changes.
          enrolled_n_ia2 <- length(enrolled_by_t)
          accrual_hat_ia2 <- if (cut_time > 0) enrolled_n_ia2 / cut_time else NA_real_
          predicted_enroll_close <- if (!is.na(accrual_hat_ia2) && accrual_hat_ia2 > 0) {
            interim_n / accrual_hat_ia2
          } else {
            NA_real_
          }

          ia2_adapt_cut_time <- if (!is.na(predicted_enroll_close)) {
            min(cut_time, predicted_enroll_close - min_months_to_close_for_adapt)
          } else {
            cut_time
          }
          ia2_adapt_cut_time <- max(ia2_adapt_cut_time, ia_times[1] + 0.5)

          enrolled_n_adapt <- sum(
            id_enroll$id %in% analysis_ids &
              id_enroll$enroll_time <= ia2_adapt_cut_time
          )
          ia2_enroll_frac <- if (interim_n > 0) enrolled_n_adapt / interim_n else NA_real_
          accrual_hat <- if (ia2_adapt_cut_time > 0) {
            enrolled_n_adapt / ia2_adapt_cut_time
          } else {
            NA_real_
          }
          ia2_months_to_close_pred <- if (!is.na(accrual_hat) && accrual_hat > 0) {
            max(0, (interim_n - enrolled_n_adapt) / accrual_hat)
          } else {
            NA_real_
          }
          ia2_adapt_allowed <- !is.na(ia2_enroll_frac) &&
            is.finite(ia2_adapt_cut_time) &&
            ia2_adapt_cut_time <= cut_time &&
            ia2_enroll_frac <= max_enrollment_frac_for_ia2
        }
      } else {
        # Final: all current_n subjects, wait for max follow-up
        analysis_ids <- all_ids[seq_len(min(current_n, length(all_ids)))]
        last_enroll <- max(id_enroll$enroll_time[id_enroll$id %in% analysis_ids])
        analysis_pool <- sim_data[sim_data$id %in% analysis_ids, ]
        # Ensure all selected subjects are followed through their observed
        # event/dropout/follow-up process by cutting at the latest observed
        # calendar time for selected participants.
        cut_time <- max(
          last_enroll + max_followup,
          max(analysis_pool$calendar_time, na.rm = TRUE)
        )
      }
      cal_times[a] <- cut_time

      class(analysis_pool) <- class(sim_data)
      cut_data <- cut_data_by_date(analysis_pool, cut_date = cut_time, event_gap = event_gap)

      if (length(unique(cut_data$id)) < 10) next

      test_res <- tryCatch(mutze_test(cut_data), error = function(e) NULL)
      if (is.null(test_res)) next

      z_val <- -test_res$z
      info_val <- 1 / test_res$se^2
      if_val <- info_val / target_info

      z_values[a] <- z_val
      info_values[a] <- info_val
      if_values[a] <- if_val

      # --- GS bounds ---
      # Always use k=2+ so gsDesign can compute interim bounds
      # Append a placeholder final analysis at timing=1 if needed
      timing_vec <- c(if_values[1:a], if (a < k_analyses) 1)
      spend_vec <- c(
        pmin(planned_timing[1:a], if_values[1:a]),
        if (a < k_analyses) 1
      )
      if (a == k_analyses) {
        timing_vec[a] <- 1
        spend_vec[a] <- 1
      }

      # Ensure monotonicity
      for (j in 2:length(timing_vec)) {
        if (!is.na(timing_vec[j]) && !is.na(timing_vec[j-1]) &&
            timing_vec[j] <= timing_vec[j-1])
          timing_vec[j] <- timing_vec[j-1] + 0.001
        if (!is.na(spend_vec[j]) && !is.na(spend_vec[j-1]) &&
            spend_vec[j] <= spend_vec[j-1])
          spend_vec[j] <- spend_vec[j-1] + 0.001
      }

      if (any(is.na(timing_vec))) next

      k_gs <- length(timing_vec)
      gs_updated <- tryCatch(
        gsDesign(
          k = k_gs, test.type = gs_plan$test.type,
          alpha = alpha_plan, beta = 1 - power_plan,
          sfu = gs_plan$upper$sf, sfupar = gs_plan$upper$param,
          sfl = gs_plan$lower$sf, sflpar = gs_plan$lower$param,
          n.fix = target_info,
          timing = timing_vec, usTime = spend_vec, lsTime = spend_vec
        ),
        error = function(e) NULL
      )
      if (is.null(gs_updated)) next

      if (z_val > gs_updated$upper$bound[a]) {
        reject <- TRUE
        reject_stage <- if (a < k_analyses) paste0("IA", a) else "Final"
        stopped <- TRUE
      } else if (a < k_analyses && !ignore_futility &&
                 if_val >= min_if_futility &&
                 z_val < gs_updated$lower$bound[a]) {
        futility_stop <- TRUE
        futility_stage <- paste0("IA", a)
        n_futility_report <- sum(
          id_enroll$id %in% analysis_ids &
            id_enroll$enroll_time <= (cut_time + analysis_lag_months)
        )
        n_futility_report <- max(n_futility_report, length(enrolled_by_t))
        n_futility_report <- min(n_futility_report, length(analysis_ids))
        n_strategies[[strat]] <- n_futility_report
        stopped <- TRUE
      }

      # --- SSR at second interim only ---
      if (a == 2 && !stopped && strat != "No adaptation" && ia2_adapt_allowed) {
        adapt_cut_data <- cut_data
        if (!is.na(ia2_adapt_cut_time)) {
          adapt_pool <- sim_data[sim_data$id %in% analysis_ids, ]
          class(adapt_pool) <- class(sim_data)
          adapt_cut_data <- cut_data_by_date(adapt_pool, cut_date = ia2_adapt_cut_time, event_gap = event_gap)
        }
        if (length(unique(adapt_cut_data$id)) < 10) next

        if (strat == "Blinded SSR") {
          blind_res <- tryCatch(
            calculate_blinded_info(adapt_cut_data, ratio = 1,
              lambda1_planning = lambda1_plan,
              lambda2_planning = lambda2_plan,
              event_gap = event_gap),
            error = function(e) NULL
          )
          if (!is.null(blind_res) && !is.na(blind_res$lambda1_adjusted)) {
            ssr_design <- tryCatch(
              sample_size_nbinom(
                lambda1 = blind_res$lambda1_adjusted,
                lambda2 = blind_res$lambda2_adjusted,
                dispersion = blind_res$dispersion_blinded,
                power = power_plan, alpha = alpha_plan,
                accrual_rate = accrual_rate_plan,
                accrual_duration = accrual_dur_plan,
                trial_duration = trial_dur_plan,
                max_followup = max_followup,
                event_gap = event_gap
              ),
              error = function(e) NULL
            )
            if (!is.null(ssr_design)) {
              n_new <- min(
                max(ceiling(ssr_design$n_total * gs_inflation), n_planned),
                n_max
              )
              ia2_adapt_applied <- (n_new > n_planned)
              n_strategies[[strat]] <- n_new
            }
          }
        } else {
          unblind_fit <- tryCatch(
            MASS::glm.nb(events ~ treatment + offset(log(tte)),
                          data = adapt_cut_data),
            error = function(e) NULL
          )
          if (!is.null(unblind_fit)) {
            k_est <- 1 / unblind_fit$theta
            lambda1_est <- exp(coef(unblind_fit)[1])
            ssr_design <- tryCatch(
              sample_size_nbinom(
                lambda1 = lambda1_est,
                lambda2 = lambda1_est * rr_plan,
                dispersion = k_est,
                power = power_plan, alpha = alpha_plan,
                accrual_rate = accrual_rate_plan,
                accrual_duration = accrual_dur_plan,
                trial_duration = trial_dur_plan,
                max_followup = max_followup,
                event_gap = event_gap
              ),
              error = function(e) NULL
            )
            if (!is.null(ssr_design)) {
              n_new <- min(
                max(ceiling(ssr_design$n_total * gs_inflation), n_planned),
                n_max
              )
              ia2_adapt_applied <- (n_new > n_planned)
              n_strategies[[strat]] <- n_new
            }
          }
        }
      }
    }

    strategy_results[[strat]] <- data.frame(
      sim = sim_id, strategy = strat,
      reject = reject, futility = futility_stop,
      reject_stage = reject_stage,
      futility_stage = futility_stage,
      n_adapted = n_strategies[[strat]],
      adapted = n_strategies[[strat]] > n_planned,
      z_ia1 = z_values[1], if_ia1 = if_values[1],
      ia1_time = cal_times[1],
      ia1_info_method = ia_info_method[1],
      ia1_fallback = ia_info_method[1] != "blinded",
      z_ia2 = z_values[2], if_ia2 = if_values[2],
      ia2_time = cal_times[2],
      ia2_info_method = ia_info_method[2],
      ia2_fallback = ia_info_method[2] != "blinded",
      ia2_enroll_frac = ia2_enroll_frac,
      ia2_months_to_close_pred = ia2_months_to_close_pred,
      ia2_adapt_cut_time = ia2_adapt_cut_time,
      ia2_adapt_allowed = ia2_adapt_allowed,
      ia2_adapt_applied = ia2_adapt_applied,
      z_final = z_values[3],
      info_final = info_values[3],
      if_final = if_values[3],
      final_time = cal_times[3],
      stringsAsFactors = FALSE
    )
  }

  do.call(rbind, strategy_results)
}
```

## Running simulations

```{r run_simulations, warning=FALSE}
precomputed_outputs <- NULL
if (use_precomputed) {
  precomputed_outputs <- readRDS(precomputed_file)
  all_results <- as.data.frame(precomputed_outputs$plot_data)
  sim_runtime_seconds <- precomputed_outputs$sim_runtime_seconds
  workers <- if (!is.null(precomputed_outputs$workers)) {
    as.integer(precomputed_outputs$workers)
  } else {
    max(1L, future::availableCores() - 1L)
  }
  sim_mode <- "Loaded precomputed outputs"
} else {
  run_scenario <- function(sc_idx) {
    sc <- scenarios[sc_idx, ]
    message(sprintf("Starting scenario %d / %d: RR=%.2f, lambda1=%.2f, k=%.2f, accrual=%.1f", 
                    sc_idx, nrow(scenarios), sc$rr_true, sc$lambda1_true, sc$k_true, sc$accrual_true))
    set.seed(1000 + sc_idx)
    results_list <- lapply(seq_len(sc$n_sims), function(sim_id) {
      tryCatch(
        run_one_trial(
          sim_id = sim_id,
          lambda1_true = sc$lambda1_true, rr_true = sc$rr_true,
          k_true = sc$k_true,
          accrual_true = sc$accrual_true,
          accrual_eff = sc$accrual_eff,
          n_planned = n_planned, n_max = n_max,
          max_followup = max_followup,
          gs_plan = gs_plan, target_info = target_info,
          planned_timing = planned_timing, gs_inflation = gs_inflation,
          lambda1_plan = lambda1_plan, lambda2_plan = lambda2_plan,
          rr_plan = rr_plan, accrual_rate_plan = accrual_rate_plan,
          accrual_dur_plan = accrual_dur_plan,
          trial_dur_plan = trial_dur_plan,
          power_plan = power_plan, alpha_plan = alpha_plan,
          min_if_futility = min_if_futility,
          max_enrollment_frac_for_ia2 = max_enrollment_frac_for_ia2,
          min_months_to_close_for_adapt = min_months_to_close_for_adapt,
          analysis_lag_months = analysis_lag_months,
          event_gap = event_gap_val,
          ignore_futility = FALSE
        ),
        error = function(e) NULL
      )
    })
    res <- do.call(rbind, results_list)
    if (is.null(res) || nrow(res) == 0) return(NULL)
    res$lambda1_true <- sc$lambda1_true
    res$k_true <- sc$k_true
    res$accrual_true <- sc$accrual_true
    res$accrual_eff <- sc$accrual_eff
    res$rr_true <- sc$rr_true
    res
  }

  sim_start <- Sys.time()
  workers <- max(1L, future::availableCores() - 1L)
  old_plan <- future::plan()
  future::plan(future::multisession, workers = workers)

  all_results <- future.apply::future_lapply(
    seq_len(nrow(scenarios)),
    run_scenario,
    future.seed = TRUE
  )

  future::plan(old_plan)
  all_results <- Filter(Negate(is.null), all_results)
  all_results <- do.call(rbind, all_results)
  sim_runtime_seconds <- as.numeric(difftime(Sys.time(), sim_start, units = "secs"))
  sim_mode <- "Fresh simulation"
}

# Backward-compatible defaults for precomputed files from older vignette versions
required_cols <- c(
  "ia2_adapt_cut_time",
  "ia2_enroll_frac", "ia2_months_to_close_pred",
  "ia2_adapt_allowed", "ia2_adapt_applied"
)
missing_cols <- setdiff(required_cols, names(all_results))
if (length(missing_cols) > 0) {
  for (nm in missing_cols) all_results[[nm]] <- NA
}

cat("Simulation mode:", sim_mode, "\n")
cat("Workers:", workers, "\n")
cat("Rows:", nrow(all_results), "\n")
if (!is.null(sim_runtime_seconds) && is.finite(sim_runtime_seconds)) {
  cat(sprintf("Simulation wall time: %.2f minutes (%.1f seconds)\n",
              sim_runtime_seconds / 60, sim_runtime_seconds))
}

# Separate RR = 1.0 non-binding futility check (ignore futility stopping)
# Run under design assumptions and with higher dispersion
null_nonbinding_n <- if (use_production) 20000 else n_sims_initial
null_k_scenarios <- c(k_plan, 1.0)

if (use_precomputed && !is.null(precomputed_outputs$null_nonbinding_summary)) {
  null_nonbinding_summary <- as.data.table(precomputed_outputs$null_nonbinding_summary)
  null_nonbinding_runtime_seconds <- precomputed_outputs$null_nonbinding_runtime_seconds
  null_nonbinding_mode <- "Loaded precomputed outputs"
} else {
  null_nonbinding_mode <- "Fresh simulation"
  null_start <- Sys.time()

  null_all <- list()
  old_plan_null <- future::plan()
  future::plan(future::multisession, workers = workers)

  for (k_null in null_k_scenarios) {
    run_null_nonbinding <- function(sim_id) {
      tryCatch(
        run_one_trial(
          sim_id = sim_id,
          lambda1_true = lambda1_plan, rr_true = 1.0, k_true = k_null,
          accrual_true = accrual_scenario_plan,
          accrual_eff = accrual_scenario_plan,
          n_planned = n_planned, n_max = n_max,
          max_followup = max_followup,
          gs_plan = gs_plan, target_info = target_info,
          planned_timing = planned_timing, gs_inflation = gs_inflation,
          lambda1_plan = lambda1_plan, lambda2_plan = lambda2_plan,
          rr_plan = rr_plan, accrual_rate_plan = accrual_rate_plan,
          accrual_dur_plan = accrual_dur_plan,
          trial_dur_plan = trial_dur_plan,
          power_plan = power_plan, alpha_plan = alpha_plan,
          min_if_futility = min_if_futility,
          max_enrollment_frac_for_ia2 = max_enrollment_frac_for_ia2,
          min_months_to_close_for_adapt = min_months_to_close_for_adapt,
          analysis_lag_months = analysis_lag_months,
          event_gap = event_gap_val,
          ignore_futility = TRUE
        ),
        error = function(e) NULL
      )
    }

    null_res <- future.apply::future_lapply(
      seq_len(null_nonbinding_n),
      run_null_nonbinding,
      future.seed = TRUE
    )
    null_res <- Filter(Negate(is.null), null_res)
    if (length(null_res) > 0) {
      nr <- as.data.table(do.call(rbind, null_res))
      nr$k_true <- k_null
      null_all[[length(null_all) + 1]] <- nr
    }
  }
  future::plan(old_plan_null)

  if (length(null_all) > 0) {
    null_dt <- rbindlist(null_all)
    null_nonbinding_summary <- null_dt[, .(
      n_sims = .N,
      type1_error = mean(reject, na.rm = TRUE),
      cross_ia1 = mean(reject_stage == "IA1", na.rm = TRUE),
      cross_ia2 = mean(reject_stage == "IA2", na.rm = TRUE),
      cross_final = mean(reject_stage == "Final", na.rm = TRUE),
      mean_n = mean(n_adapted, na.rm = TRUE),
      adapted_rate = mean(adapted, na.rm = TRUE)
    ), by = .(k_true, strategy)]
  } else {
    null_nonbinding_summary <- data.table(
      k_true = numeric(0), strategy = character(0),
      n_sims = integer(0), type1_error = numeric(0),
      cross_ia1 = numeric(0), cross_ia2 = numeric(0), cross_final = numeric(0),
      mean_n = numeric(0), adapted_rate = numeric(0)
    )
  }
  null_nonbinding_runtime_seconds <- as.numeric(
    difftime(Sys.time(), null_start, units = "secs")
  )
}

cat("RR=1 non-binding futility check mode:", null_nonbinding_mode, "\n")
cat("RR=1 non-binding replications per k:", null_nonbinding_n, "\n")
cat("k scenarios:", paste(null_k_scenarios, collapse = ", "), "\n")
```

## Results

```{r summarize}
dt <- as.data.table(all_results)

summary_dt <- dt[, .(
  n_sims = .N,
  rejection_rate = mean(reject, na.rm = TRUE),
  futility_rate = mean(futility, na.rm = TRUE),
  cross_ia1 = mean(reject_stage == "IA1", na.rm = TRUE),
  cross_ia2 = mean(reject_stage == "IA2", na.rm = TRUE),
  cross_final = mean(reject_stage == "Final", na.rm = TRUE),
  mean_n = round(mean(n_adapted, na.rm = TRUE)),
  sd_n = round(sd(n_adapted, na.rm = TRUE), 1),
  pct_adapted = round(100 * mean(adapted, na.rm = TRUE), 1),
  mean_if_ia1 = round(100 * mean(if_ia1, na.rm = TRUE), 1),
  mean_if_ia2 = round(100 * mean(if_ia2, na.rm = TRUE), 1),
  mean_if_final = round(100 * mean(if_final, na.rm = TRUE), 1),
  mean_ia1_time = round(mean(ia1_time, na.rm = TRUE), 1),
  mean_ia2_time = round(mean(ia2_time, na.rm = TRUE), 1),
  mean_ia2_adapt_cut = round(mean(ia2_adapt_cut_time, na.rm = TRUE), 1),
  mean_final_time = round(mean(final_time, na.rm = TRUE), 1),
  mean_ia2_enroll_pct = round(100 * mean(ia2_enroll_frac, na.rm = TRUE), 1),
  mean_ia2_months_to_close = round(mean(ia2_months_to_close_pred, na.rm = TRUE), 1),
  pct_ia2_adapt_allowed = round(100 * mean(ia2_adapt_allowed, na.rm = TRUE), 1),
  pct_ia2_adapt_applied = round(100 * mean(ia2_adapt_applied, na.rm = TRUE), 1),
  pct_reach_final = round(100 * mean(!is.na(if_final), na.rm = TRUE), 1),
  n_fallback_ia1 = sum(ia1_fallback, na.rm = TRUE),
  n_fallback_ia2 = sum(ia2_fallback, na.rm = TRUE),
  n_any_fallback = sum(ia1_fallback | ia2_fallback, na.rm = TRUE),
  pct_futility_ia1 = round(100 * mean(futility_stage == "IA1", na.rm = TRUE), 1),
  pct_futility_ia2 = round(100 * mean(futility_stage == "IA2", na.rm = TRUE), 1)
), by = .(lambda1_true, k_true, accrual_true, rr_true, strategy)]
```

```{r save_precomputed_outputs, eval=!use_precomputed && save_precomputed}
plot_data <- dt[, .(
  lambda1_true, k_true, accrual_true, rr_true, strategy,
  reject, futility, reject_stage, futility_stage,
  n_adapted, adapted,
  if_ia1, if_ia2, if_final,
  ia1_time, ia2_time, final_time,
  ia1_fallback, ia2_fallback,
  ia2_adapt_cut_time,
  ia2_enroll_frac, ia2_months_to_close_pred,
  ia2_adapt_allowed, ia2_adapt_applied
)]

dir.create(dirname(save_precomputed_path), recursive = TRUE, showWarnings = FALSE)
saveRDS(
  list(
    plot_data = as.data.frame(plot_data),
    sim_runtime_seconds = sim_runtime_seconds,
    workers = workers,
    null_nonbinding_summary = as.data.frame(null_nonbinding_summary),
    null_nonbinding_n = null_nonbinding_n,
    null_nonbinding_runtime_seconds = null_nonbinding_runtime_seconds,
    generated_at = as.character(Sys.time()),
    settings = list(
      n_sims_initial = n_sims_initial,
      use_production = use_production,
      design_note = design_note
    )
  ),
  save_precomputed_path
)
cat("Saved precomputed vignette outputs to:", save_precomputed_path, "\n")
```

```{r runtime_summary}
runtime_df <- data.frame(
  Metric = c("Simulation mode", "Workers", "Scenarios", "Replicates", "Rows", "Wall time (minutes)"),
  Value = c(
    sim_mode,
    as.character(workers),
    nrow(scenarios),
    sum(scenarios$n_sims),
    nrow(all_results),
    if (!is.null(sim_runtime_seconds) && is.finite(sim_runtime_seconds))
      sprintf("%.2f", sim_runtime_seconds / 60) else "NA"
  )
)

runtime_df |>
  gt() |>
  tab_header(
    title = "Simulation Runtime",
    subtitle = "Use precomputed outputs to avoid rerunning on pkgdown/CI/CRAN builds"
  )
```

```{r null_nonbinding_summary}
# Backward compatibility: precomputed outputs may lack k_true
null_df <- as.data.frame(null_nonbinding_summary)
if (!"k_true" %in% names(null_df)) null_df$k_true <- NA_real_
null_display <- null_df |>
  dplyr::transmute(
    k = k_true,
    Strategy = strategy,
    `Type I error` = round(type1_error, 4),
    `IA1` = round(cross_ia1, 4),
    `IA2` = round(cross_ia2, 4),
    `Final` = round(cross_final, 4),
    `Mean N` = round(mean_n, 1),
    `SSR applied (%)` = round(100 * adapted_rate, 1)
  )

null_display |>
  gt() |>
  tab_header(
    title = "Type I Error Under RR = 1.0 (Non-binding Futility)",
    subtitle = paste0(
      "Replications per k: ", null_nonbinding_n,
      " | ", null_nonbinding_mode,
      " | Runtime: ", round(null_nonbinding_runtime_seconds / 60, 2), " min"
    )
  ) |>
  tab_spanner(label = "Efficacy crossing at", columns = c("IA1", "IA2", "Final")) |>
  tab_footnote(
    paste("Futility stopping is ignored (non-binding) so all trials continue to",
          "the final analysis unless stopped for efficacy.",
          "'SSR applied' is the percentage of trials where the adapted N exceeded",
          "the planned N (design uses k =", k_plan, ").",
          "Under the null, SSR may still increase N because",
          "nuisance parameter estimates can differ from planning values.")
  )
```

### Power by rate ratio and SSR strategy

```{r power_curves, fig.width=10, fig.height=6}
power_avg <- dt[rr_true < 1.0, .(
  power = mean(reject, na.rm = TRUE),
  mean_final_if = mean(if_final, na.rm = TRUE),
  mean_final_month = mean(final_time, na.rm = TRUE)
), by = .(rr_true, strategy)]

ggplot(power_avg, aes(x = rr_true, y = power,
                       color = strategy, linetype = strategy)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = power_plan, linetype = "dashed", alpha = 0.4) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  scale_x_continuous(breaks = seq(0.5, 0.9, 0.1)) +
  labs(
    title = "Power by Rate Ratio and SSR Strategy",
    subtitle = paste("Averaged across nuisance scenarios |", design_note),
    x = "True Rate Ratio", y = "Power",
    color = "Strategy", linetype = "Strategy"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

```{r no_adapt_power_loss, fig.width=10, fig.height=5}
power_rr_plan <- summary_dt[
  rr_true == rr_plan &
    strategy %in% c("No adaptation", "Blinded SSR", "Unblinded SSR")
]
power_rr_plan[, strategy := factor(strategy,
  levels = c("No adaptation", "Blinded SSR", "Unblinded SSR"))]
power_rr_plan[, k_label := paste0("k = ", k_true)]
power_rr_plan[, accrual_label := paste0(accrual_true, " pts/mo")]

ggplot(power_rr_plan,
       aes(x = lambda1_true, y = rejection_rate,
           color = strategy, shape = strategy)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = power_plan, linetype = "dashed", alpha = 0.4) +
  facet_grid(k_label ~ accrual_label) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2),
                     labels = scales::percent) +
  labs(
    title = sprintf("Power at Planned RR (%.1f) by Nuisance Scenario", rr_plan),
    subtitle = paste("Dashed line = target power", scales::percent(power_plan),
                     "|", design_note),
    x = expression(lambda[1]~(true)), y = "Power",
    color = "Strategy", shape = "Strategy"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

### Power by nuisance scenario {.tabset}

```{r power_by_nuisance_tabs, results='asis', fig.width=13, fig.height=10, echo=FALSE}
rr_values <- sort(unique(dt$rr_true[dt$rr_true <= 1]))

for (rr_val in rr_values) {
  cat(sprintf("\n\n#### RR = %s\n\n", rr_val))

  stage_dt <- dt[rr_true == rr_val, .(
    `Cross IA1` = mean(reject_stage == "IA1", na.rm = TRUE),
    `Cross IA2` = mean(reject_stage == "IA2", na.rm = TRUE),
    `Cross Final` = mean(reject_stage == "Final", na.rm = TRUE),
    `No cross Final` = mean(reject_stage == "No reject" & !futility, na.rm = TRUE),
    `Futility IA2` = mean(futility_stage == "IA2", na.rm = TRUE),
    `Futility IA1` = mean(futility_stage == "IA1", na.rm = TRUE)
  ), by = .(lambda1_true, k_true, accrual_true, strategy)]

  stage_long <- data.table::melt(
    stage_dt,
    id.vars = c("lambda1_true", "k_true", "accrual_true", "strategy"),
    variable.name = "outcome",
    value.name = "prob"
  )
  stage_long[, outcome := factor(outcome, levels = c(
    "Cross IA1", "Cross IA2", "Cross Final",
    "No cross Final", "Futility IA2", "Futility IA1"
  ))]
  stage_long[, strategy := factor(strategy,
    levels = c("No adaptation", "Blinded SSR", "Unblinded SSR"))]
  stage_long[, lambda1_label := paste0("lambda1 = ", lambda1_true)]
  stage_long[, k_label := paste0("k = ", k_true)]
  stage_long[, accrual_label := paste0("accrual = ", accrual_true, "/mo")]

  if (nrow(stage_long) == 0) next

  p <- ggplot(stage_long, aes(x = strategy, y = prob, fill = outcome)) +
    geom_col(width = 0.75, color = "white", linewidth = 0.2,
             position = position_stack(reverse = TRUE)) +
    facet_grid(k_label + accrual_label ~ lambda1_label) +
    scale_fill_manual(
      values = c(
        "Cross IA1" = "#9ecae1", "Cross IA2" = "#4292c6", "Cross Final" = "#084594",
        "No cross Final" = "#fc9272", "Futility IA2" = "#de2d26", "Futility IA1" = "#a50f15"
      )
    ) +
    scale_y_continuous(breaks = seq(0, 1, 0.2), labels = scales::percent_format()) +
    labs(
      title = sprintf("Trial Outcomes by Analysis Stage (RR = %s)", rr_val),
      subtitle = design_note,
      x = "Strategy", y = "Probability", fill = "Outcome"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 20, hjust = 1))
  print(p)
  cat("\n\n")
}
```

### Adapted sample size distribution {.tabset}

```{r adapted_n_tabs, results='asis', fig.width=13, fig.height=10, echo=FALSE}
GeomHalfViolin <- ggplot2::ggproto(
  "GeomHalfViolin",
  ggplot2::GeomViolin,
  draw_group = function(data, ..., side = "l", draw_quantiles = NULL) {
    data <- transform(
      data,
      xminv = x - violinwidth * (x - xmin),
      xmaxv = x + violinwidth * (xmax - x)
    )
    newdata <- if (side == "l") {
      transform(data, x = xminv)[order(data$y), ]
    } else {
      transform(data, x = xmaxv)[order(data$y, decreasing = TRUE), ]
    }
    newdata <- rbind(newdata[1, ], newdata, newdata[nrow(newdata), ], newdata[1, ])
    ggplot2:::ggname(
      "geom_half_violin",
      ggplot2::GeomPolygon$draw_panel(newdata, ...)
    )
  }
)

geom_half_violin <- function(mapping = NULL, data = NULL, stat = "ydensity",
                             position = "identity", ..., side = c("l", "r"),
                             trim = TRUE, scale = "area", na.rm = FALSE,
                             show.legend = NA, inherit.aes = TRUE) {
  side <- match.arg(side)
  ggplot2::layer(
    data = data,
    mapping = mapping,
    stat = stat,
    geom = GeomHalfViolin,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(side = side, trim = trim, scale = scale, na.rm = na.rm, ...)
  )
}

for (rr_val in rr_values) {
  cat(sprintf("\n\n#### RR = %s\n\n", rr_val))

  violin_dt <- dt[rr_true == rr_val & strategy %in% c("Blinded SSR", "Unblinded SSR")]
  violin_dt[, strategy := factor(strategy, levels = c("Unblinded SSR", "Blinded SSR"))]
  violin_dt[, lambda1_label := paste0("lambda1 = ", lambda1_true)]
  violin_dt[, k_label := paste0("k = ", k_true)]
  violin_dt[, accrual_label := factor(accrual_true)]

  p <- ggplot(violin_dt, aes(x = accrual_label, y = n_adapted)) +
    geom_half_violin(
      data = violin_dt[strategy == "Unblinded SSR"],
      aes(fill = strategy, group = interaction(accrual_label, strategy)),
      side = "l", alpha = 0.7, scale = "width", trim = FALSE, color = NA
    ) +
    geom_half_violin(
      data = violin_dt[strategy == "Blinded SSR"],
      aes(fill = strategy, group = interaction(accrual_label, strategy)),
      side = "r", alpha = 0.7, scale = "width", trim = FALSE, color = NA
    ) +
    geom_hline(yintercept = n_planned, linetype = "dashed", alpha = 0.5) +
    geom_hline(yintercept = n_max, linetype = "dotted", color = "red",
               alpha = 0.5) +
    scale_fill_manual(
      values = c("Unblinded SSR" = "#084594", "Blinded SSR" = "#9ecae1"),
      breaks = c("Unblinded SSR", "Blinded SSR")
    ) +
    facet_grid(k_label ~ lambda1_label) +
    scale_y_continuous(breaks = seq(0, n_max + 50, 50)) +
    labs(
      title = sprintf("Adapted Sample Size at RR = %s", rr_val),
      subtitle = sprintf("Split violins: Unblinded (left) vs Blinded (right) SSR | Dashed=planned N (%d), Red dotted=cap (%d) | %s",
                          n_planned, n_max, design_note),
      x = "Accrual Rate (pts/month)", y = "Adapted N", fill = "Strategy"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
  print(p)
  cat("\n\n")
}
```


### Expected sample size at study stop {.tabset}

```{r expected_n_tabs, results='asis', fig.width=10, fig.height=7, echo=FALSE}
for (rr_val in rr_values) {
  cat(sprintf("\n\n#### RR = %s\n\n", rr_val))

  en_dt <- dt[rr_true == rr_val]
  en_dt[, strategy := factor(strategy,
    levels = c("No adaptation", "Blinded SSR", "Unblinded SSR"))]
  en_dt[, lambda1_label := paste0("lambda1 = ", lambda1_true)]
  en_dt[, k_label := paste0("k = ", k_true)]

  en_summary <- en_dt[, .(
    mean_n = mean(n_adapted, na.rm = TRUE),
    sd_n = sd(n_adapted, na.rm = TRUE)
  ), by = .(lambda1_true, k_true, accrual_true, strategy,
            lambda1_label, k_label)]

  p <- ggplot(en_summary,
    aes(x = factor(accrual_true), y = mean_n,
        fill = strategy)) +
    geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.85) +
    geom_hline(yintercept = n_planned, linetype = "dashed", alpha = 0.5) +
    facet_grid(k_label ~ lambda1_label) +
    labs(
      title = sprintf("Expected Sample Size at Study Stop (RR = %s)", rr_val),
      subtitle = paste("Dashed = planned N;",
        "interim stops count enrollment through cutoff + 2 months |",
        design_note),
      x = "Accrual rate (pts/mo)", y = "Mean sample size", fill = "Strategy"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
  print(p)
  cat("\n\n")
}
```

### Calendar and information at each analysis

The plots below use split (side-by-side) violin distributions for
$k = 0.5$ and $k = 1.0$, with vertical panels for IA1, IA2, and
Final analysis. Panels use free y-scales.

```{r analysis_time_violin, fig.width=12, fig.height=10}
time_long <- dt[strategy == "No adaptation" & rr_true == rr_plan, .(
  lambda1_true, k_true, accrual_true,
  IA1 = ia1_time, IA2 = ia2_time, Final = final_time
)]
time_long <- data.table::melt(
  time_long,
  id.vars = c("lambda1_true", "k_true", "accrual_true"),
  variable.name = "analysis",
  value.name = "calendar_time"
)
time_long[, analysis := factor(analysis, levels = c("IA1", "IA2", "Final"))]
time_long[, lambda1_label := paste0("lambda1 = ", lambda1_true)]
time_long[, k_label := paste0("k = ", k_true)]

planned_time_df <- data.frame(
  analysis = factor(c("IA1", "IA2", "Final"), levels = c("IA1", "IA2", "Final")),
  planned_time = c(analysis_times_plan[1], analysis_times_plan[2], analysis_times_plan[3])
)

ggplot(time_long,
       aes(x = factor(accrual_true), y = calendar_time, fill = factor(k_true))) +
  geom_violin(position = position_dodge(width = 0.85),
              alpha = 0.7, scale = "width", trim = FALSE) +
  geom_hline(
    data = planned_time_df,
    aes(yintercept = planned_time),
    linetype = "dashed", color = "darkgreen", inherit.aes = FALSE
  ) +
  facet_grid(analysis ~ lambda1_label, scales = "free_y") +
  scale_fill_manual(values = c("0.5" = "#6BAED6", "1" = "#2171B5")) +
  labs(
    title = "Calendar Time Distribution by Analysis (RR = 0.7, No adaptation)",
    subtitle = paste("Dashed green = planned analysis time |", design_note),
    x = "Accrual rate (pts/month)",
    y = "Calendar month",
    fill = "k"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r analysis_info_violin, fig.width=12, fig.height=10}
if_long <- dt[strategy == "No adaptation" & rr_true == rr_plan, .(
  lambda1_true, k_true, accrual_true,
  IA1 = 100 * if_ia1, IA2 = 100 * if_ia2, Final = 100 * if_final
)]
if_long <- data.table::melt(
  if_long,
  id.vars = c("lambda1_true", "k_true", "accrual_true"),
  variable.name = "analysis",
  value.name = "info_fraction"
)
if_long[, analysis := factor(analysis, levels = c("IA1", "IA2", "Final"))]
if_long[, lambda1_label := paste0("lambda1 = ", lambda1_true)]

planned_if_df <- data.frame(
  analysis = factor(c("IA1", "IA2", "Final"), levels = c("IA1", "IA2", "Final")),
  planned_if = 100 * c(planned_timing[1], planned_timing[2], 1)
)

ggplot(if_long,
       aes(x = factor(accrual_true), y = info_fraction, fill = factor(k_true))) +
  geom_violin(position = position_dodge(width = 0.85),
              alpha = 0.7, scale = "width", trim = FALSE) +
  geom_hline(
    data = planned_if_df,
    aes(yintercept = planned_if),
    linetype = "dashed", color = "darkgreen", inherit.aes = FALSE
  ) +
  facet_grid(analysis ~ lambda1_label, scales = "free_y") +
  scale_fill_manual(values = c("0.5" = "#6BAED6", "1" = "#2171B5")) +
  labs(
    title = "Information Fraction Distribution by Analysis (RR = 0.7, No adaptation)",
    subtitle = paste("Dashed green = planned information fraction |", design_note),
    x = "Accrual rate (pts/month)",
    y = "Information fraction (%)",
    fill = "k"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Discussion

### Key findings

1. **Type I error control.** Under the null (RR $\geq$ 1.0), all strategies
   maintain the one-sided Type I error near or below 2.5%. For RR = 1.0,
   futility stopping is disabled to document Type I error under the
   non-binding futility rule.

2. **Largest no-adaptation power deficit.** At planned RR = 0.7, the largest
   deficit from the 90% target under no adaptation is concentrated in
   scenarios with lower event rates and larger dispersion (see the
   "Power Shortfall Without Adaptation" table).

3. **IA2-only adaptation recovers power where information is low.** In lower
   event-rate / higher-dispersion scenarios, IA2 adaptation raises final
   information and materially improves power versus no adaptation, while still
   preserving IA1-only monitoring (see "Power by nuisance scenario" and
   "Final information at final analysis by strategy").

### Information-based interim timing

By triggering interims when blinded information reaches the target fraction
(rather than at fixed calendar times), the information fractions are
stabilized across scenarios. This prevents the anomaly where high event rates
cause excessive information at a fixed calendar time, leading to overspending
or premature decisions.

In this update, IA2 remains information-based while adaptation uses a
lead-time cutoff of at least
`r min_months_to_close_for_adapt` months before predicted enrollment close.

### Futility at low information

Futility assessment is deferred when the observed information fraction is below
`r 100 * min_if_futility`%. The spending function approach uses
$\text{usTime} = \text{lsTime} = \min(t_{\text{planned}}, t_{\text{actual}})$
to cap spending.

IA1 includes efficacy/futility monitoring but does not permit SSR.
SSR is attempted only at IA2. For operational feasibility, SSR uses an
adaptation cutoff at
$\min(\text{IA2 time},\; \text{predicted enrollment close} -
`r min_months_to_close_for_adapt` \text{ months})$, with enrollment fraction at or
below `r 100 * max_enrollment_frac_for_ia2`%.

When a trial stops early for futility, the reported final sample size includes
subjects enrolled by the stop analysis date plus
`r analysis_lag_months` months to reflect enrollment that continues during
analysis/decision implementation.

Futility and efficacy crossing percentages are reported separately for IA1 and
IA2 in the main simulation table.

### Blinded information fallback

At each candidate cut time the simulation first attempts to estimate
blinded statistical information via `calculate_blinded_info()`.
If the blinded estimate is unavailable or non-positive (e.g., too few events
for the NB model to converge), the simulation falls back to unblinded
information from `mutze_test()` ($\mathcal{I} = 1/\text{SE}^2$).
If both methods fail, a hard-coded constant of 100 is used as a last-resort
placeholder so that the search does not stall.
The simulation results table reports the number of trials that required
this fallback at each interim analysis.

### Computational considerations

Production-scale runs (e.g., 3 600 replicates per scenario for power
evaluation and 20 000 replicates under the null for Type I error)
are computationally intensive.
Parallel execution via the `future` / `future.apply` framework is
essential.
The simulation runtime table above reports the wall-clock time and
number of workers used for the current build.
For reference, a production run on a Mac Studio with an M4 chip using
11 of 12 available cores completed in approximately 2--3 hours.

## References
