---
title: "SSR simulation study"
output: rmarkdown::html_vignette
bibliography: gsDesignNB.bib
vignette: >
  %\VignetteIndexEntry{SSR simulation study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 7
)
design_note <- "Design summary computed below"
```

```{r libraries, message=FALSE, warning=FALSE}
library(gsDesignNB)
library(gsDesign)
library(data.table)
library(MASS)
library(ggplot2)
library(dplyr)
library(gt)
library(DT)
library(future)
library(future.apply)
```

## Introduction

This vignette presents a simulation study evaluating **sample size
re-estimation (SSR)** for group sequential trials with negative binomial
recurrent event endpoints. We compare three strategies:

1. **No adaptation** -- the trial proceeds with the planned sample size.
2. **Blinded SSR** [@friede2010blinded] -- nuisance parameters are
   re-estimated from pooled (treatment-blinded) interim data.
3. **Unblinded SSR** -- nuisance parameters are re-estimated from
   treatment-specific interim data.

For hypothesis testing, all interim and final analyses use the
`mutze_test()` Wald statistic. This test underpins both efficacy/futility
crossing decisions and the unblinded-information fallback used when blinded
information estimation fails.

Interim analyses are triggered when the **blinded statistical information**
reaches the planned fraction of the target, rather than at fixed calendar
times. This ensures analyses occur at comparable information levels
regardless of the true nuisance parameters.

## Planned trial design

```{r planned_parameters}
lambda1_plan      <- 0.5
rr_plan           <- 0.7
lambda2_plan      <- lambda1_plan * rr_plan
k_plan            <- 0.3
power_plan        <- 0.9
alpha_plan        <- 0.025
accrual_rate_plan <- 30
accrual_scenario_plan <- 18
accrual_dur_plan  <- 12
max_followup      <- 12
trial_dur_plan    <- accrual_dur_plan + max_followup
analysis_times_plan <- c(9, 14, 24)  # Calendar times under plan
```

### Fixed design

```{r fixed_design}
design_plan <- sample_size_nbinom(
  lambda1 = lambda1_plan, lambda2 = lambda2_plan,
  dispersion = k_plan, power = power_plan, alpha = alpha_plan,
  accrual_rate = accrual_rate_plan,
  accrual_duration = accrual_dur_plan,
  trial_duration = trial_dur_plan,
  max_followup = max_followup
)
cat("Fixed-design sample size:", design_plan$n_total, "\n")
```

### Group sequential design

We use a 3-analysis group sequential design with HSD efficacy spending and
Cauchy futility spending calibrated so the futility bounds correspond to
approximately RR = 0.9 at each analysis.

Specifically, `sfl = sfCauchy` with `sflpar = c(0.4, 0.75, 0.56, 0.63)` gives
planned lower bounds that are close to declaring futility when the observed RR
is greater than about 0.9 at either interim. Because the design uses
non-binding futility (`test.type = 4`), ignoring a lower-bound crossing does
not invalidate the final efficacy decision, which is still based on upper
boundary control.

```{r gs_design}
gs_plan <- design_plan |>
  gsNBCalendar(
    k = 3, test.type = 4, alpha = alpha_plan,
    sfu = sfHSD, sfupar = -2,
    sfl = sfCauchy, sflpar = c(0.4, 0.75, 0.56, 0.63),
    analysis_times = analysis_times_plan
  ) |>
  gsDesignNB::toInteger()

n_planned <- gs_plan$n_total[gs_plan$k]
target_info <- gs_plan$n.I[gs_plan$k]
planned_timing <- gs_plan$timing
gs_inflation <- n_planned / design_plan$n_total
accrual_rate_plan_eff <- n_planned / accrual_dur_plan
design_note <- paste0(
  "Design: lambda1=", lambda1_plan,
  ", RR=", rr_plan,
  ", k=", k_plan,
  ", planned accrual=", round(accrual_rate_plan_eff, 1), "/mo",
  ", planned N=", n_planned,
  ", max follow-up=", max_followup, " mo"
)

cat("Planned N (GS):", n_planned, "\n")
cat("Information fractions:", round(planned_timing, 3), "\n")
cat("GS inflation factor:", round(gs_inflation, 3), "\n")
cat("Effective accrual rate for GS design:", round(accrual_rate_plan_eff, 2), "/month\n")
```

```{r gs_bound_summary}
gsBoundSummary(gs_plan,
  deltaname = "RR", logdelta = TRUE,
  Nname = "Information", timename = "Month",
  digits = 4, ddigits = 2
) |>
  as.data.frame() |>
  gt() |>
  tab_header(
    title = "Planned Group Sequential Design",
    subtitle = design_note
  ) |>
  tab_footnote(
    "Cauchy futility spending gives planned futility near observed RR > 0.9 at IA1/IA2; lower bounds are non-binding."
  ) |>
  tab_footnote(
    footnote = sprintf(
      "Planned cumulative sample size: IA1 = %.0f, IA2 = %.0f, Final = %.0f.",
      gs_plan$n_total[1], gs_plan$n_total[2], gs_plan$n_total[3]
    )
  )
```

### Expected information fraction at planned time of each interim

This table shows the expected information fraction at each planned interim
calendar time under each nuisance scenario. Under the design assumptions
(bold green row), the information fractions match the GS design
(`r paste(round(100 * planned_timing[1:2], 1), collapse = "%, ")`%).

```{r info_fractions}
nuisance_grid <- expand.grid(
  lambda1_true = c(0.3, 0.5, 0.8),
  k_true = c(0.3, 0.6),
  accrual_true = c(12, 18, 24),
  stringsAsFactors = FALSE
)

# Accrual scenarios use effective monthly enrollment rates directly.

for (a in 1:2) {
  col_name <- paste0("IF_analysis_", a)
  nuisance_grid[[col_name]] <- sapply(seq_len(nrow(nuisance_grid)), function(i) {
    accrual_eff <- nuisance_grid$accrual_true[i]
    info_at_t <- compute_info_at_time(
      analysis_time = analysis_times_plan[a],
      accrual_rate = accrual_eff,
      accrual_duration = accrual_dur_plan,
      lambda1 = nuisance_grid$lambda1_true[i],
      lambda2 = nuisance_grid$lambda1_true[i] * rr_plan,
      dispersion = nuisance_grid$k_true[i],
      max_followup = max_followup
    )
    round(100 * info_at_t / target_info, 1)
  })
}

nuisance_grid$is_planned <- with(nuisance_grid,
  lambda1_true == 0.5 & k_true == 0.3 & accrual_true == accrual_scenario_plan)

nuisance_grid |>
  select(-is_planned) |>
  gt() |>
  tab_header(
    title = "Expected Information Fraction (%) at Planned Time of Each Interim",
    subtitle = design_note
  ) |>
  cols_label(
    lambda1_true = "lambda1 (true)",
    k_true = "k (true)",
    accrual_true = "Accrual (pts/mo)",
    IF_analysis_1 = paste0("IA 1 (mo ", analysis_times_plan[1], ")"),
    IF_analysis_2 = paste0("IA 2 (mo ", analysis_times_plan[2], ")")
  ) |>
  tab_style(
    style = list(cell_fill(color = "#E8F5E9"), cell_text(weight = "bold")),
    locations = cells_body(rows = nuisance_grid$is_planned)
  ) |>
  tab_footnote(
    footnote = paste(
      "Computed via compute_info_at_time() divided by planned final information.",
      "Accrual values (12/18/24) are effective enrollment rates used directly.",
      "Bold green = design assumptions.",
      "With information-based timing, interims occur when blinded info reaches",
      "the target fraction, so the calendar time varies by scenario."
    )
  )
```

## Scenario grid

```{r scenario_grid}
scenarios <- expand.grid(
  lambda1_true = c(0.3, 0.5, 0.8),
  k_true = c(0.3, 0.6),
  accrual_true = c(12, 18, 24),
  rr_true = c(0.5, 0.6, 0.7, 0.8, 0.9, 1.0, 1.1),
  stringsAsFactors = FALSE
)

n_sims_initial <- 50
use_production <- identical(Sys.getenv("GSDESIGNNB_PRODUCTION_SSR"), "true")

scenarios$n_sims <- ifelse(
  use_production,
  3600,
  n_sims_initial
)
scenarios$accrual_eff <- scenarios$accrual_true

n_max <- 2 * n_planned
min_if_futility <- 0.3
target_if <- planned_timing  # Target IF for each analysis
# IA2 adaptation gate (less strict than prior 80% / 2 months setting)
max_enrollment_frac_for_ia2 <- 1.00
min_months_to_close_for_adapt <- 2
analysis_lag_months <- 2

# Optional precomputed outputs for fast vignette builds
precomputed_basename <- "ssr_sim_vignette_outputs.rds"
project_root <- if (file.exists("DESCRIPTION")) "." else
  if (file.exists("../DESCRIPTION")) ".." else "."
precomputed_source_path <- file.path(
  project_root, "inst", "extdata", precomputed_basename
)
precomputed_file <- system.file("extdata", precomputed_basename, package = "gsDesignNB")
if (precomputed_file == "" && file.exists(precomputed_source_path)) {
  precomputed_file <- precomputed_source_path
}
force_rerun <- identical(Sys.getenv("GSDESIGNNB_FORCE_SSR_SIM"), "true")
use_precomputed <- (!use_production) && !force_rerun && nzchar(precomputed_file)
save_precomputed <- identical(Sys.getenv("GSDESIGNNB_SAVE_SSR_OUTPUTS"), "true")
save_precomputed_path <- precomputed_source_path

cat("Scenarios:", nrow(scenarios), "| Replicates:", sum(scenarios$n_sims), "\n")
cat("Accrual rates used in simulation:", paste(round(sort(unique(scenarios$accrual_true)), 1), collapse = ", "),
    "/month\n")
cat(
  "IA2 SSR gate: adaptation uses cutoff at min(IA2 time, predicted close - ",
  min_months_to_close_for_adapt,
  " months); enrollment cap <= ",
  100 * max_enrollment_frac_for_ia2,
  "%.\n",
  sep = ""
)
cat("Futility-stop sample size counts enrollment through +",
    analysis_lag_months, " months after stop.\n", sep = "")
cat("Production plan: 3600 reps for each main-grid scenario;\n")
cat("separate RR=1 non-binding futility check uses 20000 reps.\n")
if (use_precomputed) cat("Using precomputed outputs:", precomputed_file, "\n")
```

## Simulation engine

Interim analyses are information-based and use dynamic spending:

1. **IA1 (~40% IF)**: efficacy/futility only; no SSR adaptation.
2. **IA2 (target ~76% IF)**: efficacy/futility evaluated using observed IF
   and spending.
3. **SSR at IA2 only**: nuisance estimates for adaptation use an IA2
   adaptation cutoff time at
   $\min(\text{IA2 time}, \text{predicted enrollment close} -
   `r min_months_to_close_for_adapt` months), enforcing at least
   `r min_months_to_close_for_adapt` months operational lead-time.
   The enrollment-fraction cap is set in the scenario setup.

At each analysis, bounds are recalculated from observed information and
spending time. If blinded NB information fails, timing falls back to
`mutze_test()` information (tracked in output).

```{r simulation_engine}
run_one_trial <- function(sim_id, lambda1_true, rr_true, k_true,
                          accrual_true, accrual_eff, n_planned, n_max,
                          max_followup, gs_plan, target_info,
                          planned_timing, gs_inflation,
                          lambda1_plan, lambda2_plan, rr_plan,
                          accrual_rate_plan, accrual_dur_plan,
                          trial_dur_plan, power_plan, alpha_plan,
                          min_if_futility,
                          max_enrollment_frac_for_ia2,
                          min_months_to_close_for_adapt,
                          analysis_lag_months,
                          ignore_futility = FALSE) {
  lambda2_true <- lambda1_true * rr_true
  k_analyses <- 3

  enroll_rate_sim <- data.frame(
    rate = accrual_eff, duration = n_max / accrual_eff
  )
  fail_rate_sim <- data.frame(
    treatment = c("Control", "Experimental"),
    rate = c(lambda1_true, lambda2_true),
    dispersion = k_true
  )
  dropout_rate_sim <- data.frame(
    treatment = c("Control", "Experimental"),
    rate = c(0, 0), duration = c(100, 100)
  )
  sim_data <- nb_sim(
    enroll_rate = enroll_rate_sim, fail_rate = fail_rate_sim,
    dropout_rate = dropout_rate_sim, max_followup = max_followup,
    n = n_max
  )

  id_enroll <- unique(sim_data[, c("id", "enroll_time")])
  id_enroll <- id_enroll[order(id_enroll$enroll_time, id_enroll$id), ]
  all_ids <- id_enroll$id
  max_calendar <- max(sim_data$calendar_time) + 1

  n_strategies <- c(
    "No adaptation" = n_planned,
    "Blinded SSR" = n_planned,
    "Unblinded SSR" = n_planned
  )

  # Helper: info and method at a given cut time
  info_method_at_t <- function(pool, t) {
    cut_dt <- cut_data_by_date(pool, cut_date = t)
    if (sum(cut_dt$events) < 2 || length(unique(cut_dt$id)) < 4) {
      return(list(info = 0, method = "insufficient"))
    }

    blind_res <- tryCatch(
      calculate_blinded_info(
        cut_dt, ratio = 1,
        lambda1_planning = lambda1_plan,
        lambda2_planning = lambda2_plan
      ),
      error = function(e) NULL
    )
    if (!is.null(blind_res) &&
        is.finite(blind_res$blinded_info) &&
        !is.na(blind_res$blinded_info) &&
        blind_res$blinded_info > 0) {
      return(list(info = blind_res$blinded_info, method = "blinded"))
    }

    mutze_res <- tryCatch(mutze_test(cut_dt), error = function(e) NULL)
    if (!is.null(mutze_res) &&
        is.finite(mutze_res$se) &&
        !is.na(mutze_res$se) &&
        mutze_res$se > 0) {
      return(list(info = 1 / mutze_res$se^2, method = "unblinded"))
    }

    # User-requested last-resort fallback
    list(info = 100, method = "fallback100")
  }

  # --- Find information-based interim times ---
  # Restrict to planned N for interim timing search
  planned_pool <- sim_data[sim_data$id %in% all_ids[seq_len(n_planned)], ]
  class(planned_pool) <- class(sim_data)

  ia_times <- numeric(k_analyses)
  ia_info_method <- rep("blinded", k_analyses)

  for (a in 1:(k_analyses - 1)) {
    target_info_a <- planned_timing[a] * target_info
    min_t <- if (a == 1) 1 else ia_times[a - 1] + 0.5

    ia_times[a] <- tryCatch(
      get_cut_date(
        data = planned_pool,
        target_info = target_info_a,
        lambda1 = lambda1_plan, lambda2 = lambda2_plan,
        ratio = 1, min_date = min_t, max_date = max_calendar
      ),
      error = function(e) max_calendar
    )

    # Track whether blinded or fallback method was needed at this cut
    ia_info_method[a] <- info_method_at_t(planned_pool, ia_times[a])$method
  }

  # --- Run each strategy ---
  strategy_results <- list()

  for (strat in names(n_strategies)) {
    stopped <- FALSE
    reject <- FALSE
    futility_stop <- FALSE
    reject_stage <- "No reject"
    futility_stage <- "No futility"
    ia2_enroll_frac <- NA_real_
    ia2_months_to_close_pred <- NA_real_
    ia2_adapt_allowed <- FALSE
    ia2_adapt_applied <- FALSE
    ia2_adapt_cut_time <- NA_real_
    n_futility_report <- NA_real_
    z_values <- rep(NA_real_, k_analyses)
    info_values <- rep(NA_real_, k_analyses)
    if_values <- rep(NA_real_, k_analyses)
    cal_times <- rep(NA_real_, k_analyses)

    for (a in seq_len(k_analyses)) {
      if (stopped) break

      current_n <- n_strategies[[strat]]

      if (a < k_analyses) {
        # Interim analyses are based on planned N for all strategies.
        # SSR changes the final sample size target after IA2.
        cut_time <- ia_times[a]
        interim_n <- min(n_planned, length(all_ids))
        analysis_ids <- all_ids[seq_len(interim_n)]
        enrolled_by_t <- id_enroll$id[
          id_enroll$id %in% analysis_ids & id_enroll$enroll_time <= cut_time
        ]
        analysis_pool <- sim_data[sim_data$id %in% enrolled_by_t, ]

        if (a == 2) {
          # Adaptation cutoff may be earlier than IA2 efficacy/futility
          # to ensure time to implement enrollment changes.
          enrolled_n_ia2 <- length(enrolled_by_t)
          accrual_hat_ia2 <- if (cut_time > 0) enrolled_n_ia2 / cut_time else NA_real_
          predicted_enroll_close <- if (!is.na(accrual_hat_ia2) && accrual_hat_ia2 > 0) {
            interim_n / accrual_hat_ia2
          } else {
            NA_real_
          }

          ia2_adapt_cut_time <- if (!is.na(predicted_enroll_close)) {
            min(cut_time, predicted_enroll_close - min_months_to_close_for_adapt)
          } else {
            cut_time
          }
          ia2_adapt_cut_time <- max(ia2_adapt_cut_time, ia_times[1] + 0.5)

          enrolled_n_adapt <- sum(
            id_enroll$id %in% analysis_ids &
              id_enroll$enroll_time <= ia2_adapt_cut_time
          )
          ia2_enroll_frac <- if (interim_n > 0) enrolled_n_adapt / interim_n else NA_real_
          accrual_hat <- if (ia2_adapt_cut_time > 0) {
            enrolled_n_adapt / ia2_adapt_cut_time
          } else {
            NA_real_
          }
          ia2_months_to_close_pred <- if (!is.na(accrual_hat) && accrual_hat > 0) {
            max(0, (interim_n - enrolled_n_adapt) / accrual_hat)
          } else {
            NA_real_
          }
          ia2_adapt_allowed <- !is.na(ia2_enroll_frac) &&
            is.finite(ia2_adapt_cut_time) &&
            ia2_adapt_cut_time <= cut_time &&
            ia2_enroll_frac <= max_enrollment_frac_for_ia2
        }
      } else {
        # Final: all current_n subjects, wait for max follow-up
        analysis_ids <- all_ids[seq_len(min(current_n, length(all_ids)))]
        last_enroll <- max(id_enroll$enroll_time[id_enroll$id %in% analysis_ids])
        analysis_pool <- sim_data[sim_data$id %in% analysis_ids, ]
        # Ensure all selected subjects are followed through their observed
        # event/dropout/follow-up process by cutting at the latest observed
        # calendar time for selected participants.
        cut_time <- max(
          last_enroll + max_followup,
          max(analysis_pool$calendar_time, na.rm = TRUE)
        )
      }
      cal_times[a] <- cut_time

      class(analysis_pool) <- class(sim_data)
      cut_data <- cut_data_by_date(analysis_pool, cut_date = cut_time)

      if (length(unique(cut_data$id)) < 10) next

      test_res <- tryCatch(mutze_test(cut_data), error = function(e) NULL)
      if (is.null(test_res)) next

      z_val <- -test_res$z
      info_val <- 1 / test_res$se^2
      if_val <- info_val / target_info

      z_values[a] <- z_val
      info_values[a] <- info_val
      if_values[a] <- if_val

      # --- GS bounds ---
      # Always use k=2+ so gsDesign can compute interim bounds
      # Append a placeholder final analysis at timing=1 if needed
      timing_vec <- c(if_values[1:a], if (a < k_analyses) 1)
      spend_vec <- c(
        pmin(planned_timing[1:a], if_values[1:a]),
        if (a < k_analyses) 1
      )
      if (a == k_analyses) {
        timing_vec[a] <- 1
        spend_vec[a] <- 1
      }

      # Ensure monotonicity
      for (j in 2:length(timing_vec)) {
        if (!is.na(timing_vec[j]) && !is.na(timing_vec[j-1]) &&
            timing_vec[j] <= timing_vec[j-1])
          timing_vec[j] <- timing_vec[j-1] + 0.001
        if (!is.na(spend_vec[j]) && !is.na(spend_vec[j-1]) &&
            spend_vec[j] <= spend_vec[j-1])
          spend_vec[j] <- spend_vec[j-1] + 0.001
      }

      if (any(is.na(timing_vec))) next

      k_gs <- length(timing_vec)
      gs_updated <- tryCatch(
        gsDesign(
          k = k_gs, test.type = gs_plan$test.type,
          alpha = alpha_plan, beta = 1 - power_plan,
          sfu = gs_plan$upper$sf, sfupar = gs_plan$upper$param,
          sfl = gs_plan$lower$sf, sflpar = gs_plan$lower$param,
          n.fix = target_info,
          timing = timing_vec, usTime = spend_vec, lsTime = spend_vec
        ),
        error = function(e) NULL
      )
      if (is.null(gs_updated)) next

      if (z_val > gs_updated$upper$bound[a]) {
        reject <- TRUE
        reject_stage <- if (a < k_analyses) paste0("IA", a) else "Final"
        stopped <- TRUE
      } else if (a < k_analyses && !ignore_futility &&
                 if_val >= min_if_futility &&
                 z_val < gs_updated$lower$bound[a]) {
        futility_stop <- TRUE
        futility_stage <- paste0("IA", a)
        n_futility_report <- sum(
          id_enroll$id %in% analysis_ids &
            id_enroll$enroll_time <= (cut_time + analysis_lag_months)
        )
        n_futility_report <- max(n_futility_report, length(enrolled_by_t))
        n_futility_report <- min(n_futility_report, length(analysis_ids))
        n_strategies[[strat]] <- n_futility_report
        stopped <- TRUE
      }

      # --- SSR at second interim only ---
      if (a == 2 && !stopped && strat != "No adaptation" && ia2_adapt_allowed) {
        adapt_cut_data <- cut_data
        if (!is.na(ia2_adapt_cut_time)) {
          adapt_pool <- sim_data[sim_data$id %in% analysis_ids, ]
          class(adapt_pool) <- class(sim_data)
          adapt_cut_data <- cut_data_by_date(adapt_pool, cut_date = ia2_adapt_cut_time)
        }
        if (length(unique(adapt_cut_data$id)) < 10) next

        if (strat == "Blinded SSR") {
          blind_res <- tryCatch(
            calculate_blinded_info(adapt_cut_data, ratio = 1,
              lambda1_planning = lambda1_plan,
              lambda2_planning = lambda2_plan),
            error = function(e) NULL
          )
          if (!is.null(blind_res) && !is.na(blind_res$lambda1_adjusted)) {
            ssr_design <- tryCatch(
              sample_size_nbinom(
                lambda1 = blind_res$lambda1_adjusted,
                lambda2 = blind_res$lambda2_adjusted,
                dispersion = blind_res$dispersion_blinded,
                power = power_plan, alpha = alpha_plan,
                accrual_rate = accrual_rate_plan,
                accrual_duration = accrual_dur_plan,
                trial_duration = trial_dur_plan,
                max_followup = max_followup
              ),
              error = function(e) NULL
            )
            if (!is.null(ssr_design)) {
              n_new <- min(
                max(ceiling(ssr_design$n_total * gs_inflation), n_planned),
                n_max
              )
              ia2_adapt_applied <- (n_new > n_planned)
              n_strategies[[strat]] <- n_new
            }
          }
        } else {
          unblind_fit <- tryCatch(
            MASS::glm.nb(events ~ treatment + offset(log(tte)),
                          data = adapt_cut_data),
            error = function(e) NULL
          )
          if (!is.null(unblind_fit)) {
            k_est <- 1 / unblind_fit$theta
            lambda1_est <- exp(coef(unblind_fit)[1])
            ssr_design <- tryCatch(
              sample_size_nbinom(
                lambda1 = lambda1_est,
                lambda2 = lambda1_est * rr_plan,
                dispersion = k_est,
                power = power_plan, alpha = alpha_plan,
                accrual_rate = accrual_rate_plan,
                accrual_duration = accrual_dur_plan,
                trial_duration = trial_dur_plan,
                max_followup = max_followup
              ),
              error = function(e) NULL
            )
            if (!is.null(ssr_design)) {
              n_new <- min(
                max(ceiling(ssr_design$n_total * gs_inflation), n_planned),
                n_max
              )
              ia2_adapt_applied <- (n_new > n_planned)
              n_strategies[[strat]] <- n_new
            }
          }
        }
      }
    }

    strategy_results[[strat]] <- data.frame(
      sim = sim_id, strategy = strat,
      reject = reject, futility = futility_stop,
      reject_stage = reject_stage,
      futility_stage = futility_stage,
      n_adapted = n_strategies[[strat]],
      adapted = n_strategies[[strat]] > n_planned,
      z_ia1 = z_values[1], if_ia1 = if_values[1],
      ia1_time = cal_times[1],
      ia1_info_method = ia_info_method[1],
      ia1_fallback = ia_info_method[1] != "blinded",
      z_ia2 = z_values[2], if_ia2 = if_values[2],
      ia2_time = cal_times[2],
      ia2_info_method = ia_info_method[2],
      ia2_fallback = ia_info_method[2] != "blinded",
      ia2_enroll_frac = ia2_enroll_frac,
      ia2_months_to_close_pred = ia2_months_to_close_pred,
      ia2_adapt_cut_time = ia2_adapt_cut_time,
      ia2_adapt_allowed = ia2_adapt_allowed,
      ia2_adapt_applied = ia2_adapt_applied,
      z_final = z_values[3],
      info_final = info_values[3],
      if_final = if_values[3],
      final_time = cal_times[3],
      stringsAsFactors = FALSE
    )
  }

  do.call(rbind, strategy_results)
}
```

## Running simulations

```{r run_simulations, warning=FALSE}
precomputed_outputs <- NULL
if (use_precomputed) {
  precomputed_outputs <- readRDS(precomputed_file)
  all_results <- as.data.frame(precomputed_outputs$plot_data)
  sim_runtime_seconds <- precomputed_outputs$sim_runtime_seconds
  workers <- if (!is.null(precomputed_outputs$workers)) {
    as.integer(precomputed_outputs$workers)
  } else {
    max(1L, future::availableCores() - 1L)
  }
  sim_mode <- "Loaded precomputed outputs"
} else {
  run_scenario <- function(sc_idx) {
    sc <- scenarios[sc_idx, ]
    message(sprintf("Starting scenario %d / %d: RR=%.2f, lambda1=%.2f, k=%.2f, accrual=%.1f", 
                    sc_idx, nrow(scenarios), sc$rr_true, sc$lambda1_true, sc$k_true, sc$accrual_true))
    set.seed(1000 + sc_idx)
    results_list <- lapply(seq_len(sc$n_sims), function(sim_id) {
      tryCatch(
        run_one_trial(
          sim_id = sim_id,
          lambda1_true = sc$lambda1_true, rr_true = sc$rr_true,
          k_true = sc$k_true,
          accrual_true = sc$accrual_true,
          accrual_eff = sc$accrual_eff,
          n_planned = n_planned, n_max = n_max,
          max_followup = max_followup,
          gs_plan = gs_plan, target_info = target_info,
          planned_timing = planned_timing, gs_inflation = gs_inflation,
          lambda1_plan = lambda1_plan, lambda2_plan = lambda2_plan,
          rr_plan = rr_plan, accrual_rate_plan = accrual_rate_plan,
          accrual_dur_plan = accrual_dur_plan,
          trial_dur_plan = trial_dur_plan,
          power_plan = power_plan, alpha_plan = alpha_plan,
          min_if_futility = min_if_futility,
          max_enrollment_frac_for_ia2 = max_enrollment_frac_for_ia2,
          min_months_to_close_for_adapt = min_months_to_close_for_adapt,
          analysis_lag_months = analysis_lag_months,
          ignore_futility = FALSE
        ),
        error = function(e) NULL
      )
    })
    res <- do.call(rbind, results_list)
    if (is.null(res) || nrow(res) == 0) return(NULL)
    res$lambda1_true <- sc$lambda1_true
    res$k_true <- sc$k_true
    res$accrual_true <- sc$accrual_true
    res$accrual_eff <- sc$accrual_eff
    res$rr_true <- sc$rr_true
    res
  }

  sim_start <- Sys.time()
  workers <- max(1L, future::availableCores() - 1L)
  old_plan <- future::plan()
  future::plan(future::multisession, workers = workers)

  all_results <- future.apply::future_lapply(
    seq_len(nrow(scenarios)),
    run_scenario,
    future.seed = TRUE
  )

  future::plan(old_plan)
  all_results <- Filter(Negate(is.null), all_results)
  all_results <- do.call(rbind, all_results)
  sim_runtime_seconds <- as.numeric(difftime(Sys.time(), sim_start, units = "secs"))
  sim_mode <- "Fresh simulation"
}

# Backward-compatible defaults for precomputed files from older vignette versions
required_cols <- c(
  "ia2_adapt_cut_time",
  "ia2_enroll_frac", "ia2_months_to_close_pred",
  "ia2_adapt_allowed", "ia2_adapt_applied"
)
missing_cols <- setdiff(required_cols, names(all_results))
if (length(missing_cols) > 0) {
  for (nm in missing_cols) all_results[[nm]] <- NA
}

cat("Simulation mode:", sim_mode, "\n")
cat("Workers:", workers, "\n")
cat("Rows:", nrow(all_results), "\n")
if (!is.null(sim_runtime_seconds) && is.finite(sim_runtime_seconds)) {
  cat(sprintf("Simulation wall time: %.2f minutes (%.1f seconds)\n",
              sim_runtime_seconds / 60, sim_runtime_seconds))
}

# Separate RR = 1.0 non-binding futility check (ignore futility stopping)
null_nonbinding_n <- if (use_production) 20000 else n_sims_initial
if (use_precomputed && !is.null(precomputed_outputs$null_nonbinding_summary)) {
  null_nonbinding_summary <- as.data.table(precomputed_outputs$null_nonbinding_summary)
  null_nonbinding_runtime_seconds <- precomputed_outputs$null_nonbinding_runtime_seconds
  null_nonbinding_mode <- "Loaded precomputed outputs"
} else {
  null_nonbinding_mode <- "Fresh simulation"
  null_start <- Sys.time()

  run_null_nonbinding <- function(sim_id) {
    if (sim_id %% 1000 == 0) message(sprintf("Processing RR=1 null simulation %d / %d", sim_id, null_nonbinding_n))
    tryCatch(
      run_one_trial(
        sim_id = sim_id,
        lambda1_true = lambda1_plan, rr_true = 1.0, k_true = k_plan,
        accrual_true = accrual_scenario_plan,
        accrual_eff = accrual_scenario_plan,
        n_planned = n_planned, n_max = n_max,
        max_followup = max_followup,
        gs_plan = gs_plan, target_info = target_info,
        planned_timing = planned_timing, gs_inflation = gs_inflation,
        lambda1_plan = lambda1_plan, lambda2_plan = lambda2_plan,
        rr_plan = rr_plan, accrual_rate_plan = accrual_rate_plan,
        accrual_dur_plan = accrual_dur_plan,
        trial_dur_plan = trial_dur_plan,
        power_plan = power_plan, alpha_plan = alpha_plan,
        min_if_futility = min_if_futility,
        max_enrollment_frac_for_ia2 = max_enrollment_frac_for_ia2,
        min_months_to_close_for_adapt = min_months_to_close_for_adapt,
        analysis_lag_months = analysis_lag_months,
        ignore_futility = TRUE
      ),
      error = function(e) NULL
    )
  }

  old_plan_null <- future::plan()
  future::plan(future::multisession, workers = workers)
  null_results <- future.apply::future_lapply(
    seq_len(null_nonbinding_n),
    run_null_nonbinding,
    future.seed = TRUE
  )
  future::plan(old_plan_null)

  null_results <- Filter(Negate(is.null), null_results)
  if (length(null_results) > 0) {
    null_dt <- as.data.table(do.call(rbind, null_results))
    null_nonbinding_summary <- null_dt[, .(
      n_sims = .N,
      type1_error = mean(reject, na.rm = TRUE),
      cross_ia1 = mean(reject_stage == "IA1", na.rm = TRUE),
      cross_ia2 = mean(reject_stage == "IA2", na.rm = TRUE),
      cross_final = mean(reject_stage == "Final", na.rm = TRUE),
      futility_rate = mean(futility, na.rm = TRUE),
      futility_ia1 = mean(futility_stage == "IA1", na.rm = TRUE),
      futility_ia2 = mean(futility_stage == "IA2", na.rm = TRUE),
      mean_n = mean(n_adapted, na.rm = TRUE),
      adapted_rate = mean(adapted, na.rm = TRUE)
    ), by = .(strategy)]
  } else {
    null_nonbinding_summary <- data.table(
      strategy = character(0), n_sims = integer(0), type1_error = numeric(0),
      cross_ia1 = numeric(0), cross_ia2 = numeric(0), cross_final = numeric(0),
      futility_rate = numeric(0), futility_ia1 = numeric(0),
      futility_ia2 = numeric(0), mean_n = numeric(0), adapted_rate = numeric(0)
    )
  }
  null_nonbinding_runtime_seconds <- as.numeric(
    difftime(Sys.time(), null_start, units = "secs")
  )
}

null_required_cols <- c("futility_ia1", "futility_ia2")
null_missing <- setdiff(null_required_cols, names(null_nonbinding_summary))
if (length(null_missing) > 0) {
  for (nm in null_missing) null_nonbinding_summary[[nm]] <- NA_real_
}

cat("RR=1 non-binding futility check mode:", null_nonbinding_mode, "\n")
cat("RR=1 non-binding replications:", null_nonbinding_n, "\n")
```

## Results

```{r summarize}
dt <- as.data.table(all_results)

summary_dt <- dt[, .(
  n_sims = .N,
  rejection_rate = mean(reject, na.rm = TRUE),
  futility_rate = mean(futility, na.rm = TRUE),
  cross_ia1 = mean(reject_stage == "IA1", na.rm = TRUE),
  cross_ia2 = mean(reject_stage == "IA2", na.rm = TRUE),
  cross_final = mean(reject_stage == "Final", na.rm = TRUE),
  mean_n = round(mean(n_adapted, na.rm = TRUE)),
  sd_n = round(sd(n_adapted, na.rm = TRUE), 1),
  pct_adapted = round(100 * mean(adapted, na.rm = TRUE), 1),
  mean_if_ia1 = round(100 * mean(if_ia1, na.rm = TRUE), 1),
  mean_if_ia2 = round(100 * mean(if_ia2, na.rm = TRUE), 1),
  mean_if_final = round(100 * mean(if_final, na.rm = TRUE), 1),
  mean_ia1_time = round(mean(ia1_time, na.rm = TRUE), 1),
  mean_ia2_time = round(mean(ia2_time, na.rm = TRUE), 1),
  mean_ia2_adapt_cut = round(mean(ia2_adapt_cut_time, na.rm = TRUE), 1),
  mean_final_time = round(mean(final_time, na.rm = TRUE), 1),
  mean_ia2_enroll_pct = round(100 * mean(ia2_enroll_frac, na.rm = TRUE), 1),
  mean_ia2_months_to_close = round(mean(ia2_months_to_close_pred, na.rm = TRUE), 1),
  pct_ia2_adapt_allowed = round(100 * mean(ia2_adapt_allowed, na.rm = TRUE), 1),
  pct_ia2_adapt_applied = round(100 * mean(ia2_adapt_applied, na.rm = TRUE), 1),
  pct_reach_final = round(100 * mean(!is.na(if_final), na.rm = TRUE), 1),
  pct_futility_deferred = round(100 * mean(if_ia1 < min_if_futility, na.rm = TRUE), 1),
  pct_fallback_ia1 = round(100 * mean(ia1_fallback, na.rm = TRUE), 1),
  pct_fallback_ia2 = round(100 * mean(ia2_fallback, na.rm = TRUE), 1),
  pct_any_fallback = round(100 * mean(ia1_fallback | ia2_fallback, na.rm = TRUE), 1),
  pct_futility_ia1 = round(100 * mean(futility_stage == "IA1", na.rm = TRUE), 1),
  pct_futility_ia2 = round(100 * mean(futility_stage == "IA2", na.rm = TRUE), 1)
), by = .(lambda1_true, k_true, accrual_true, rr_true, strategy)]
```

```{r save_precomputed_outputs, eval=!use_precomputed && save_precomputed}
plot_data <- dt[, .(
  lambda1_true, k_true, accrual_true, rr_true, strategy,
  reject, futility, reject_stage, futility_stage,
  n_adapted, adapted,
  if_ia1, if_ia2, if_final,
  ia1_time, ia2_time, final_time,
  ia1_fallback, ia2_fallback,
  ia2_adapt_cut_time,
  ia2_enroll_frac, ia2_months_to_close_pred,
  ia2_adapt_allowed, ia2_adapt_applied
)]

dir.create(dirname(save_precomputed_path), recursive = TRUE, showWarnings = FALSE)
saveRDS(
  list(
    plot_data = as.data.frame(plot_data),
    sim_runtime_seconds = sim_runtime_seconds,
    workers = workers,
    null_nonbinding_summary = as.data.frame(null_nonbinding_summary),
    null_nonbinding_n = null_nonbinding_n,
    null_nonbinding_runtime_seconds = null_nonbinding_runtime_seconds,
    generated_at = as.character(Sys.time()),
    settings = list(
      n_sims_initial = n_sims_initial,
      use_production = use_production,
      design_note = design_note
    )
  ),
  save_precomputed_path
)
cat("Saved precomputed vignette outputs to:", save_precomputed_path, "\n")
```

### Simulation results

Use the column filters to focus on specific parameter combinations or rate
ratios. RR $\geq$ 1 rows show Type I error; RR < 1 rows show power.
Efficacy crossing percentages and futility stopping percentages are reported
separately for IA1 and IA2.

```{r combined_table}
display_dt <- summary_dt[, .(
  lambda1 = lambda1_true, k = k_true, Accrual = accrual_true,
  RR = rr_true, Strategy = strategy,
  Rejection = rejection_rate,
  Futility = futility_rate,
  `Futility IA1 (%)` = pct_futility_ia1,
  `Futility IA2 (%)` = pct_futility_ia2,
  `Cross IA1` = cross_ia1,
  `Cross IA2` = cross_ia2,
  `Cross Final` = cross_final,
  `Mean N` = mean_n,
  `SD N` = sd_n,
  `Adapted (%)` = pct_adapted,
  `IA2 enroll (%)` = mean_ia2_enroll_pct,
  `IA2 months to close` = mean_ia2_months_to_close,
  `IA2 adapt allowed (%)` = pct_ia2_adapt_allowed,
  `IA2 adapt applied (%)` = pct_ia2_adapt_applied,
  `IF at IA1 (%)` = mean_if_ia1,
  `IF at IA2 (%)` = mean_if_ia2,
  `IF at Final (%)` = mean_if_final,
  `IA1 month` = mean_ia1_time,
  `IA2 month` = mean_ia2_time,
  `IA2 adapt cut month` = mean_ia2_adapt_cut,
  `Final month` = mean_final_time,
  `Reach final (%)` = pct_reach_final,
  `Fut. defer (%)` = pct_futility_deferred,
  `Fallback IA1 (%)` = pct_fallback_ia1,
  `Fallback IA2 (%)` = pct_fallback_ia2,
  `Any fallback (%)` = pct_any_fallback,
  n_sims = n_sims
)]

DT::datatable(
  as.data.frame(display_dt),
  filter = "top",
  options = list(pageLength = 20, autoWidth = TRUE, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "font-size: 14px; font-weight: bold;",
    "Simulation Results | ", design_note
  )
) |>
  DT::formatPercentage(
    c(
      "Rejection", "Futility", "Cross IA1", "Cross IA2", "Cross Final",
      "Adapted (%)", "IA2 enroll (%)", "IA2 adapt allowed (%)",
      "IA2 adapt applied (%)", "IF at IA1 (%)", "IF at IA2 (%)",
      "IF at Final (%)", "Reach final (%)", "Fut. defer (%)",
      "Fallback IA1 (%)", "Fallback IA2 (%)", "Any fallback (%)",
      "Futility IA1 (%)", "Futility IA2 (%)"
    ),
    digits = 1
  ) |>
  DT::formatStyle(
    "Rejection",
    backgroundColor = DT::styleInterval(0.03, c("white", "#FFCCCC")),
    target = "cell"
  )
```

```{r stage_rates_summary}
stage_rates <- summary_dt[, .(
  `Rejection (%)` = 100 * mean(rejection_rate),
  `Futility IA1 (%)` = mean(pct_futility_ia1),
  `Futility IA2 (%)` = mean(pct_futility_ia2),
  `Cross IA1 (%)` = 100 * mean(cross_ia1),
  `Cross IA2 (%)` = 100 * mean(cross_ia2),
  `Cross Final (%)` = 100 * mean(cross_final)
), by = .(RR = rr_true, Strategy = strategy)]

stage_rates |>
  as.data.frame() |>
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 1))) |>
  gt() |>
  tab_header(
    title = "Efficacy/Futility Stage Rates by RR and Strategy",
    subtitle = "IA1 and IA2 rates shown separately"
  )
```

```{r runtime_summary}
runtime_df <- data.frame(
  Metric = c("Simulation mode", "Workers", "Scenarios", "Replicates", "Rows", "Wall time (minutes)"),
  Value = c(
    sim_mode,
    as.character(workers),
    nrow(scenarios),
    sum(scenarios$n_sims),
    nrow(all_results),
    if (!is.null(sim_runtime_seconds) && is.finite(sim_runtime_seconds))
      sprintf("%.2f", sim_runtime_seconds / 60) else "NA"
  )
)

runtime_df |>
  gt() |>
  tab_header(
    title = "Simulation Runtime",
    subtitle = "Use precomputed outputs to avoid rerunning on pkgdown/CI/CRAN builds"
  )
```

```{r null_nonbinding_summary}
null_nonbinding_summary |>
  as.data.frame() |>
  dplyr::mutate(
    type1_error = round(type1_error, 4),
    cross_ia1 = round(cross_ia1, 4),
    cross_ia2 = round(cross_ia2, 4),
    cross_final = round(cross_final, 4),
    futility_rate = round(futility_rate, 4),
    futility_ia1 = round(futility_ia1, 4),
    futility_ia2 = round(futility_ia2, 4),
    mean_n = round(mean_n, 1),
    adapted_rate = round(adapted_rate, 4)
  ) |>
  gt() |>
  tab_header(
    title = "RR = 1.0 Non-binding Futility Check (Ignored Futility Stopping)",
    subtitle = paste0(
      "Replications: ", null_nonbinding_n,
      " | Mode: ", null_nonbinding_mode,
      " | Runtime (min): ", round(null_nonbinding_runtime_seconds / 60, 2)
    )
  )
```

### Power by rate ratio and SSR strategy

```{r power_curves, fig.width=10, fig.height=6}
power_avg <- dt[rr_true < 1.0, .(
  power = mean(reject, na.rm = TRUE),
  mean_final_if = mean(if_final, na.rm = TRUE),
  mean_final_month = mean(final_time, na.rm = TRUE)
), by = .(rr_true, strategy)]

ggplot(power_avg, aes(x = rr_true, y = power,
                       color = strategy, linetype = strategy)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = power_plan, linetype = "dashed", alpha = 0.4) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  scale_x_continuous(breaks = seq(0.5, 0.9, 0.1)) +
  labs(
    title = "Power by Rate Ratio and SSR Strategy",
    subtitle = paste("Averaged across nuisance scenarios |", design_note),
    x = "True Rate Ratio", y = "Power",
    color = "Strategy", linetype = "Strategy"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

```{r no_adapt_power_loss}
power_rr_plan <- summary_dt[
  rr_true == rr_plan &
    strategy %in% c("No adaptation", "Blinded SSR", "Unblinded SSR"),
  .(lambda1_true, k_true, accrual_true, strategy, rejection_rate)
]

power_rr_plan <- dcast(
  power_rr_plan,
  lambda1_true + k_true + accrual_true ~ strategy,
  value.var = "rejection_rate"
)
power_rr_plan[, `Best SSR power` := pmax(`Blinded SSR`, `Unblinded SSR`)]
power_rr_plan[, `Best SSR method` := ifelse(`Blinded SSR` >= `Unblinded SSR`,
                                            "Blinded", "Unblinded")]
power_rr_plan[, `Power gain from best SSR` := `Best SSR power` - `No adaptation`]
power_rr_plan[, `No adaptation shortfall to 90%` := pmax(0, power_plan - `No adaptation`)]
power_rr_plan[, `Best SSR shortfall to 90%` := pmax(0, power_plan - `Best SSR power`)]
power_rr_plan <- power_rr_plan[order(-`No adaptation shortfall to 90%`)]

power_rr_display <- power_rr_plan[, .(
  lambda1_true, k_true, accrual_true,
  `No adaptation power` = `No adaptation`,
  `Blinded SSR power` = `Blinded SSR`,
  `Unblinded SSR power` = `Unblinded SSR`,
  `Best SSR power`,
  `Best SSR method`,
  `Power gain from best SSR`,
  `No adaptation shortfall to 90%`,
  `Best SSR shortfall to 90%`
)]

power_rr_display |>
  as.data.frame() |>
  dplyr::mutate(
    across(where(is.numeric), ~ round(.x, 3))
  ) |>
  gt() |>
  tab_header(
    title = "Power Shortfall Without Adaptation at Planned RR (0.7)",
    subtitle = "Includes strategy-specific power, best SSR gain, and residual shortfall"
  )
```

```{r final_metrics}
final_metrics_all <- dt[rr_true <= 1.0, .(
  Scenario = "Average over nuisance scenarios",
  `Power` = mean(reject, na.rm = TRUE),
  `Mean final IF (%)` = 100 * mean(if_final, na.rm = TRUE),
  `Mean final month` = mean(final_time, na.rm = TRUE),
  `Futility stop (%)` = 100 * mean(futility, na.rm = TRUE),
  `Adapted (%)` = 100 * mean(adapted, na.rm = TRUE)
), by = .(rr_true, strategy)]

final_metrics_design <- dt[
  rr_true <= 1.0 &
    lambda1_true == lambda1_plan &
    k_true == k_plan &
    accrual_true == accrual_scenario_plan,
  .(
    Scenario = "Design assumptions",
    `Power` = mean(reject, na.rm = TRUE),
    `Mean final IF (%)` = 100 * mean(if_final, na.rm = TRUE),
    `Mean final month` = mean(final_time, na.rm = TRUE),
    `Futility stop (%)` = 100 * mean(futility, na.rm = TRUE),
    `Adapted (%)` = 100 * mean(adapted, na.rm = TRUE)
  ),
  by = .(rr_true, strategy)
]

final_metrics <- rbind(final_metrics_all, final_metrics_design)

final_metrics |>
  as.data.frame() |>
  dplyr::arrange(rr_true, strategy, desc(Scenario)) |>
  dplyr::mutate(
    `Power` = round(`Power`, 3),
    `Mean final IF (%)` = round(`Mean final IF (%)`, 1),
    `Mean final month` = round(`Mean final month`, 1),
    `Futility stop (%)` = round(`Futility stop (%)`, 1),
    `Adapted (%)` = round(`Adapted (%)`, 1)
  ) |>
  gt() |>
  tab_header(
    title = "Final Analysis Diagnostics by RR and Strategy",
    subtitle = design_note
  ) |>
  tab_style(
    style = list(cell_fill(color = "#E8F5E9"), cell_text(weight = "bold")),
    locations = cells_body(rows = Scenario == "Design assumptions")
  )
```

```{r adapted_n_dispersion}
adapt_disp <- summary_dt[
  rr_true < 1 & strategy %in% c("Blinded SSR", "Unblinded SSR"),
  .(
    `Mean SD(adapted N)` = mean(sd_n, na.rm = TRUE),
    `Median SD(adapted N)` = median(sd_n, na.rm = TRUE),
    `Mean adapted (%)` = mean(pct_adapted, na.rm = TRUE),
    `IA2 adaptation allowed (%)` = mean(pct_ia2_adapt_allowed, na.rm = TRUE),
    `IA2 adaptation applied (%)` = mean(pct_ia2_adapt_applied, na.rm = TRUE)
  ),
  by = .(strategy)
]

adapt_disp |>
  as.data.frame() |>
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 1))) |>
  gt() |>
  tab_header(
    title = "Adapted Sample Size Dispersion (RR < 1)",
    subtitle = "Delayed IA2-only adaptation with enrollment-window eligibility"
  )
```

### Power by nuisance scenario {.tabset}

```{r power_by_nuisance_tabs, results='asis', fig.width=13, fig.height=12}
rr_values <- sort(unique(scenarios$rr_true[scenarios$rr_true <= 1]))

for (rr_val in rr_values) {
  cat(sprintf("\n\n#### RR = %s\n\n", rr_val))

  stage_dt <- dt[rr_true == rr_val, .(
    IA1 = mean(reject_stage == "IA1", na.rm = TRUE),
    IA2 = mean(reject_stage == "IA2", na.rm = TRUE),
    Final = mean(reject_stage == "Final", na.rm = TRUE)
  ), by = .(lambda1_true, k_true, accrual_true, strategy)]

  stage_long <- data.table::melt(
    stage_dt,
    id.vars = c("lambda1_true", "k_true", "accrual_true", "strategy"),
    variable.name = "analysis_stage",
    value.name = "prob"
  )
  stage_long[, analysis_stage := factor(
    analysis_stage,
    levels = c("IA1", "IA2", "Final"),
    ordered = TRUE
  )]
  stage_long[, strategy := factor(
    strategy,
    levels = c("No adaptation", "Blinded SSR", "Unblinded SSR")
  )]
  stage_long[, lambda1_label := paste0("lambda1 = ", lambda1_true)]
  stage_long[, k_label := paste0("k = ", k_true)]
  stage_long[, accrual_label := paste0("accrual = ", accrual_true, "/mo")]

  p <- ggplot(stage_long, aes(x = strategy, y = prob, fill = analysis_stage)) +
    geom_col(
      width = 0.75, color = "white", linewidth = 0.2,
      position = position_stack(reverse = TRUE)
    ) +
    facet_grid(k_label + accrual_label ~ lambda1_label) +
    scale_fill_manual(
      values = c("IA1" = "#9ecae1", "IA2" = "#4292c6", "Final" = "#084594"),
      breaks = c("IA1", "IA2", "Final"),
      limits = c("IA1", "IA2", "Final")
    ) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2),
                       labels = scales::percent_format()) +
    labs(
      title = sprintf("Efficacy Crossing Probability by Analysis Stage (RR = %s)", rr_val),
      subtitle = design_note,
      x = "Strategy",
      y = "Probability of crossing efficacy bound",
      fill = "Crossing stage"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 20, hjust = 1)
    )
  print(p)
  cat("\n\n")
}
```

### Adapted sample size distribution {.tabset}

```{r adapted_n_tabs, results='asis', fig.width=13, fig.height=10}
GeomHalfViolin <- ggplot2::ggproto(
  "GeomHalfViolin",
  ggplot2::GeomViolin,
  draw_group = function(data, ..., side = "l", draw_quantiles = NULL) {
    data <- transform(
      data,
      xminv = x - violinwidth * (x - xmin),
      xmaxv = x + violinwidth * (xmax - x)
    )
    newdata <- if (side == "l") {
      transform(data, x = xminv)[order(data$y), ]
    } else {
      transform(data, x = xmaxv)[order(data$y, decreasing = TRUE), ]
    }
    newdata <- rbind(newdata[1, ], newdata, newdata[nrow(newdata), ], newdata[1, ])
    ggplot2:::ggname(
      "geom_half_violin",
      ggplot2::GeomPolygon$draw_panel(newdata, ...)
    )
  }
)

geom_half_violin <- function(mapping = NULL, data = NULL, stat = "ydensity",
                             position = "identity", ..., side = c("l", "r"),
                             trim = TRUE, scale = "area", na.rm = FALSE,
                             show.legend = NA, inherit.aes = TRUE) {
  side <- match.arg(side)
  ggplot2::layer(
    data = data,
    mapping = mapping,
    stat = stat,
    geom = GeomHalfViolin,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(side = side, trim = trim, scale = scale, na.rm = na.rm, ...)
  )
}

for (rr_val in rr_values) {
  cat(sprintf("\n\n#### RR = %s\n\n", rr_val))

  violin_dt <- dt[rr_true == rr_val & strategy %in% c("Blinded SSR", "Unblinded SSR")]
  violin_dt[, strategy := factor(strategy, levels = c("Unblinded SSR", "Blinded SSR"))]
  violin_dt[, lambda1_label := paste0("lambda1 = ", lambda1_true)]
  violin_dt[, k_label := paste0("k = ", k_true)]
  violin_dt[, accrual_label := factor(accrual_true)]

  p <- ggplot(violin_dt, aes(x = accrual_label, y = n_adapted)) +
    geom_half_violin(
      data = violin_dt[strategy == "Unblinded SSR"],
      aes(fill = strategy, group = interaction(accrual_label, strategy)),
      side = "l", alpha = 0.7, scale = "width", trim = FALSE, color = NA
    ) +
    geom_half_violin(
      data = violin_dt[strategy == "Blinded SSR"],
      aes(fill = strategy, group = interaction(accrual_label, strategy)),
      side = "r", alpha = 0.7, scale = "width", trim = FALSE, color = NA
    ) +
    geom_hline(yintercept = n_planned, linetype = "dashed", alpha = 0.5) +
    geom_hline(yintercept = n_max, linetype = "dotted", color = "red",
               alpha = 0.5) +
    scale_fill_manual(
      values = c("Unblinded SSR" = "#084594", "Blinded SSR" = "#9ecae1"),
      breaks = c("Unblinded SSR", "Blinded SSR")
    ) +
    facet_grid(k_label ~ lambda1_label) +
    scale_y_continuous(breaks = seq(0, n_max + 50, 50)) +
    labs(
      title = sprintf("Adapted Sample Size at RR = %s", rr_val),
      subtitle = sprintf("Split violins: Unblinded (left) vs Blinded (right) SSR | Dashed=planned N (%d), Red dotted=cap (%d) | %s",
                          n_planned, n_max, design_note),
      x = "Accrual Rate (pts/month)", y = "Adapted N", fill = "Strategy"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
  print(p)
  cat("\n\n")
}
```

### Final information at final analysis by strategy {.tabset}

```{r final_info_violin_tabs, results='asis', fig.width=13, fig.height=10}
rr_values_all <- sort(unique(scenarios$rr_true))

for (rr_val in rr_values_all) {
  cat(sprintf("\n\n#### RR = %s\n\n", rr_val))

  info_dt <- dt[rr_true == rr_val & !is.na(if_final)]
  if (nrow(info_dt) == 0) {
    cat("No trials reached final analysis for this RR value.\n\n")
    next
  }
  info_dt[, strategy := factor(
    strategy,
    levels = c("No adaptation", "Blinded SSR", "Unblinded SSR")
  )]
  info_dt[, lambda1_label := paste0("lambda1 = ", lambda1_true)]
  info_dt[, k_label := paste0("k = ", k_true)]

  p <- ggplot(
    info_dt,
    aes(x = factor(accrual_true), y = 100 * if_final, fill = strategy)
  ) +
    geom_violin(position = position_dodge(width = 0.85), alpha = 0.7,
                scale = "width", trim = FALSE) +
    geom_hline(yintercept = 100, linetype = "dashed", color = "darkgreen",
               alpha = 0.6) +
    facet_grid(k_label ~ lambda1_label) +
    labs(
      title = sprintf("Final Information Fraction at RR = %s", rr_val),
      subtitle = paste(
        "Among trials reaching final analysis; dashed line = 100% target |",
        design_note
      ),
      x = "Accrual Rate (pts/month)",
      y = "Final information fraction (%)",
      fill = "Strategy"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
  print(p)
  cat("\n\n")
}
```

```{r rr08_diagnostics}
rr_diag <- dt[rr_true %in% c(0.7, 0.8) & strategy != "No adaptation", .(
  `Power` = mean(reject, na.rm = TRUE),
  `Futility stop (%)` = 100 * mean(futility, na.rm = TRUE),
  `Futility IA1 (%)` = 100 * mean(futility_stage == "IA1", na.rm = TRUE),
  `Futility IA2 (%)` = 100 * mean(futility_stage == "IA2", na.rm = TRUE),
  `Adapted (%)` = 100 * mean(adapted, na.rm = TRUE),
  `Mean adapted N` = mean(n_adapted, na.rm = TRUE),
  `Mean final IF (%)` = 100 * mean(if_final, na.rm = TRUE),
  `Mean final month` = mean(final_time, na.rm = TRUE)
), by = .(rr_true, strategy)]

rr_diag |>
  as.data.frame() |>
  dplyr::mutate(
    across(where(is.numeric), ~ round(.x, 2))
  ) |>
  gt() |>
  tab_header(
    title = "Diagnostics for RR = 0.7 vs RR = 0.8",
    subtitle = "Assessing whether lower adaptation at RR = 0.8 is driven by futility stopping"
  )
```

### Calendar and information at each analysis

The plots below use split (side-by-side) violin distributions for
\eqn{k = 0.3} and \eqn{k = 0.6}, with vertical panels for IA1, IA2, and
Final analysis. Panels use free y-scales.

```{r analysis_time_violin, fig.width=12, fig.height=10}
time_long <- dt[strategy == "No adaptation" & rr_true == rr_plan, .(
  lambda1_true, k_true, accrual_true,
  IA1 = ia1_time, IA2 = ia2_time, Final = final_time
)]
time_long <- data.table::melt(
  time_long,
  id.vars = c("lambda1_true", "k_true", "accrual_true"),
  variable.name = "analysis",
  value.name = "calendar_time"
)
time_long[, analysis := factor(analysis, levels = c("IA1", "IA2", "Final"))]
time_long[, lambda1_label := paste0("lambda1 = ", lambda1_true)]
time_long[, k_label := paste0("k = ", k_true)]

planned_time_df <- data.frame(
  analysis = factor(c("IA1", "IA2", "Final"), levels = c("IA1", "IA2", "Final")),
  planned_time = c(analysis_times_plan[1], analysis_times_plan[2], analysis_times_plan[3])
)

ggplot(time_long,
       aes(x = factor(accrual_true), y = calendar_time, fill = factor(k_true))) +
  geom_violin(position = position_dodge(width = 0.85),
              alpha = 0.7, scale = "width", trim = FALSE) +
  geom_hline(
    data = planned_time_df,
    aes(yintercept = planned_time),
    linetype = "dashed", color = "darkgreen", inherit.aes = FALSE
  ) +
  facet_grid(analysis ~ lambda1_label, scales = "free_y") +
  scale_fill_manual(values = c("0.3" = "#6BAED6", "0.6" = "#2171B5")) +
  labs(
    title = "Calendar Time Distribution by Analysis (RR = 0.7, No adaptation)",
    subtitle = paste("Dashed green = planned analysis time |", design_note),
    x = "Accrual rate (pts/month)",
    y = "Calendar month",
    fill = "k"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r analysis_info_violin, fig.width=12, fig.height=10}
if_long <- dt[strategy == "No adaptation" & rr_true == rr_plan, .(
  lambda1_true, k_true, accrual_true,
  IA1 = 100 * if_ia1, IA2 = 100 * if_ia2, Final = 100 * if_final
)]
if_long <- data.table::melt(
  if_long,
  id.vars = c("lambda1_true", "k_true", "accrual_true"),
  variable.name = "analysis",
  value.name = "info_fraction"
)
if_long[, analysis := factor(analysis, levels = c("IA1", "IA2", "Final"))]
if_long[, lambda1_label := paste0("lambda1 = ", lambda1_true)]

planned_if_df <- data.frame(
  analysis = factor(c("IA1", "IA2", "Final"), levels = c("IA1", "IA2", "Final")),
  planned_if = 100 * c(planned_timing[1], planned_timing[2], 1)
)

ggplot(if_long,
       aes(x = factor(accrual_true), y = info_fraction, fill = factor(k_true))) +
  geom_violin(position = position_dodge(width = 0.85),
              alpha = 0.7, scale = "width", trim = FALSE) +
  geom_hline(
    data = planned_if_df,
    aes(yintercept = planned_if),
    linetype = "dashed", color = "darkgreen", inherit.aes = FALSE
  ) +
  facet_grid(analysis ~ lambda1_label, scales = "free_y") +
  scale_fill_manual(values = c("0.3" = "#6BAED6", "0.6" = "#2171B5")) +
  labs(
    title = "Information Fraction Distribution by Analysis (RR = 0.7, No adaptation)",
    subtitle = paste("Dashed green = planned information fraction |", design_note),
    x = "Accrual rate (pts/month)",
    y = "Information fraction (%)",
    fill = "k"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Discussion

### Key findings

1. **Type I error control.** Under the null (RR $\geq$ 1.0), all strategies
   maintain the one-sided Type I error near or below 2.5%. For RR = 1.0,
   futility stopping is disabled to document Type I error under the
   non-binding futility rule.

2. **Largest no-adaptation power deficit.** At planned RR = 0.7, the largest
   deficit from the 90% target under no adaptation is concentrated in
   scenarios with lower event rates and larger dispersion (see the
   "Power Shortfall Without Adaptation" table).

3. **IA2-only adaptation recovers power where information is low.** In lower
   event-rate / higher-dispersion scenarios, IA2 adaptation raises final
   information and materially improves power versus no adaptation, while still
   preserving IA1-only monitoring (see "Power by nuisance scenario" and
   "Final information at final analysis by strategy").

### Information-based interim timing

By triggering interims when blinded information reaches the target fraction
(rather than at fixed calendar times), the information fractions are
stabilized across scenarios. This prevents the anomaly where high event rates
cause excessive information at a fixed calendar time, leading to overspending
or premature decisions.

In this update, IA2 remains information-based while adaptation uses a
lead-time cutoff of at least
`r min_months_to_close_for_adapt` months before predicted enrollment close.

### Futility at low information

Futility assessment is deferred when the observed information fraction is below
`r 100 * min_if_futility`%. The spending function approach uses
$\text{usTime} = \text{lsTime} = \min(t_{\text{planned}}, t_{\text{actual}})$
to cap spending.

IA1 includes efficacy/futility monitoring but does not permit SSR.
SSR is attempted only at IA2. For operational feasibility, SSR uses an
adaptation cutoff at
$\min(\text{IA2 time}, \text{predicted enrollment close} -
`r min_months_to_close_for_adapt` months), with enrollment fraction at or
below `r 100 * max_enrollment_frac_for_ia2`%.

When a trial stops early for futility, the reported final sample size includes
subjects enrolled by the stop analysis date plus
`r analysis_lag_months` months to reflect enrollment that continues during
analysis/decision implementation.

Futility and efficacy crossing percentages are reported separately for IA1 and
IA2 in the main simulation table.

### Blinded information fallback

When blinded information cannot be estimated at a candidate cut time, the
timing search falls back to unblinded information from `mutze_test()`.
If both blinded and unblinded information fail, a user-requested constant
fallback value of `100` is used. Fallback frequencies are reported in the
simulation table.

## References
